---
output: 
  bookdown::pdf_document2: default
  stevetemplates::article3:
    keep_tex: false
    toc: false
    number_section: true
    fig_caption: yes
title: "Disagreement on a bench: an empirical analysis of dissent at the Czech Constitutional Court"
author:
- name: ANONYMIZED
- name: ANONYMIZED
anonymous: FALSE
removetitleabstract: FALSE
abstract: "This article examines the factors influencing dissenting opinions by judges at the Czech Constitutional Court (CCC). We build on the disagreement-identification model. Our findings do not support the existence of a strong norm of consensus operating at the CCC. However, the complexity of a case, measured by the number of legal acts involved, is positively correlated with the likelihood of a dissenting opinion. Similarly, cases concerning controversial topics are more likely to generate dissents.  A placebo test strengthens this finding by demonstrating that randomly chosen, non-controversial topics have minimal impact on dissenting behavior. When facing a high workload, judges are less likely to write separate opinions. Interestingly, the time a judge has been in office does not seem to influence their likelihood of dissenting, indicating a consistent level of importance placed on collegiality."
thanks: "Replication files are available on the author's Github account ANONYMIZED. **Current version**: `r format(Sys.time(), '%B %d, %Y')`"
keywords: "empirical legal research, courts, dissents, judicial behavior, political science, regression analysis"
date: "`r format(Sys.time(), '%B %d, %Y')`"
link-citations: yes
linkcolor: blue
geometry: margin=1in
fontsize: 11pt
doublespacing: TRUE
endnote: no
# pandocparas: TRUE
sansitup: FALSE
mainfont: Times New Roman
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{float}
bibliography: "`r rbbt::bbt_write_bib('bibliography.bib', overwrite = TRUE)`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(patchwork)
library(tidyverse); theme_set(theme_minimal())
library(ggplot2)
library(kableExtra)
library(modelsummary) 
library(bookdown)

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, fig.width=4.2, fig.asp=0.8)

load("../data/descriptive_statistics.RData")
load("../data/model_results.RData")
```

# Introduction
"I don't like them [separate opinions]. (...) Because I am a routine judge and I am of the opinion that when a collegiate body makes a decision, a person X shouldn't further comment on it, just because they were of a differing view. It is undermining the authority of that court." The previous quote has been voiced by one of the Czech Constitutional Court ("CCC") judges. The view that judges are pro-claimers of law and that a court must exercise an unanimous authority has firm roots in the European legal tradition, in which legal jurisprudence has mainly concerned itself with black letter law and its doctrine and theory. It has overlooked the internal processes and motivations and external influences on judges. Without intending to delve into the long history of scholarship on the nature of law, legal interpretation and the role of judges in the whole process, legal philosopher Hans Kelsen pronounced that "the pure theory of law does not concern itself with internal mental nor with material processes." [@kelsenReineRechtslehreEinleitung1934].

Such a view complicates the study of a dissenting voice. A separate opinion shows lack of unity at the court; does that not undermine its role and authority as a pro-claimer of law? How could we explain all the variation in the dissenting behavior of judges just by taking into account law? It is not to say that such a view is wide shared, it is not, it is rather to show that the political, strategical, or external influences and considerations on judges have in the European legal scholarship not yet received and enjoyed the same degree of attention as in the US context. There, the empirical legal scholarship has flourished and has produced work on whether and to what extent they behave as for example political or strategic actors [@carrubbaWhoControlsContent2012; @clarkLocatingSupremeCourt2010; @epsteinChoicesJusticesMake1997; @lauderdaleScalingPoliticallyMeaningful2014; @sunsteinAreJudgesPolitical2006; @cameronChapterWhatJudges2017; @clarkEstimatingEffectLeisure2018; @epsteinWhyWhenJudges2011; @epsteinStrategicRevolutionJudicial2000; @kornhauserModelingCollegialCourts1992; @kornhauserModelingCollegialCourts1992a; @posnerWhatJudgesJustices1993; @posnerHowJudgesThink2010; @rousseyOverburdenedJudges2018].

That is not without its own reasons. The US institutional setting, especially at the SCOTUS, gives judges more room to channel their preferences. The judges enjoy a life time tenure, the nomination procedure enjoy attention and is conducted in a more political manner, SCOTUS has control over its docket and separate opinions have had a long history to build upon. In contrast to that, in some European countries, the judges of the apex courts are not even permitted to attach separate opinions [for a complete overview, see @kelemenJudicialDissentEuropean2017]. Others, including Czechia, have permitted them. Allowing for separate opinions gives judges plenty of leeway to not only simply express their competing legal point of view but also to push through their political, strategical, or other (such as leisure) preferences or to simply channel their frustration. To the extent that other influences other than simple black letter law come into play in the decision whether and how to dissent, the study of separate opinion is a study of individual judicial decision-making in general.

Regarding the puzzle of separate opinions and disagreement on the bench in the US context, @epsteinWhyWhenJudges2011 have come up with a theoretical model of dissenting behavior and dissent aversion and they tested to what extent do US judges behave strategically when attaching separate opinions. They found that the US judges are influenced by their political and strategic consideration, such as the collegiality costs a separate opinion may bring about or the workload they currently have. In the European context, @kelemenJudicialDissentEuropean2017 offers a mainly theoretical comparative overview of the various regimes of dissenting behavior across European courts. Hanretty has made use of dissenting behavior of Spanish and Portuguese judges to conduct a point estimation of location of the "political" position of the judges [@hanrettyDissentIberiaIdeal2012] or to investigate what does a dissent reveal about the dimension across which the disagreement runs on Estonian Supreme Court [@hanrettyJudicialDisagreementNeed2015] and the British Law Lords [@hanrettyDecisionsIdealPoints2013]. Most importantly, @wittigOccurrenceSeparateOpinions2016 has put forward and empirically tested on the German Federal Constitutional Court a theoretical *disagreement-identification* model of dissenting behavior tailored at the European context that aims to overcome the shortcomings of the traditional accounts of judicial behavior stemming from the US context. 

In the Czech context, legal scholarship concerning separate opinions has been sparse, empirical legal scholarship even sparser. The latter has focused mainly on analyzing the presence of implicit voting blocs (called coalitions) in the last term of the CCC. Because the votes are not known, in conducting a network analysis, the researches have relied on the information, which judge dissented in which decision. The research has revealed that the third period of CCC between 2013-2023 is rather polarized and that there are two big dissenting coalitions of judges that clash against each other [@chmelCoOvlivnujeUstavni2021; @smekalMimopravniVlivyNa2021; @vartazaryanSitOvaAnalyza2022]. The @smekalMimopravniVlivyNa2021 book goes so far to coin the first coalition as a more left-leaning and the second as a more right-leaning. We are not convinced by this label.

There are multiple shortcomings of these studies that leave a research gap for us to explore. First, the aforementioned studies apart from the @epsteinWhyWhenJudges2011 and @wittigOccurrenceSeparateOpinions2016 focus on the implications of a dissent rather than its circumstances and its causes. That includes the studies on the CCC. Second, the Czech studies include in their analysis only cases with a separate opinion and do not compare it against decisions without it. To this end, we believe that the studies suffer from selection bias and lack robustness. Does the coalition theory also hold outside the context of the plenary decisions with a separate opinion? Moreover, they rely primarily on network analysis of the dissenting opinions in the plenary proceedings and support their claims with interviews with the judges. Employment of quantitative methods to seek out trends is warranted. Lastly, the theories of (individual) judicial decision-making have been for the most part developed for and have been tested in the US context, apart from the @wittigOccurrenceSeparateOpinions2016 thesis. The pressing question remains to what extent do the conclusions made in a completely different context carry over elsewhere. We do not consider the transfer to be self-evident due to an incomparable political, institutional and legal setting.

We believe the CCC to be an object worthy of study. First, concerning the institutional setting,^[We delve into the institutional design of the CCC deeper in the following section.] the appointment procedure of the CCC judges may be compared to the SCOTUS, whereas its role within the constitutional system is akin to the European constitutional courts, with the German federal constitutional court at its forefront. The CCC has been vested with large amount of competences: it may review laws in abstract as well as in concrete proceedings initiated by an individual complaint.  On top of that, the CCC has broadened its power and may now review even constitutional amendments and historically has not been afraid to step in into politically salient cases. The decisions of the CCC have also enjoyed political and media attention: in the *Melčák* case, the CCC entered a heated political battlefield by annulling a constitutional amendment, which shortened the term of the Chamber of Deputies, and sparking a drawn out constitutional crisis.^[The English version of the decision is available at: https://www.usoud.cz/en/decisions/2009-09-10-pl-us-27-09-constitutional-act-on-shortening-the-term-of-office-of-the-chamber-of-deputies.] Secondly, separate opinions both in the form of a dissenting opinion as well as a concurring opinion have been allowed for at the CCC since its inception. Therefore, a dissent has had a long history at the CCC and the practice of it is now firmly established. Lastly, thanks to the CCC Database [@paulikCzechConstitutionalCourt2024] all the data required to conduct empirical legal research are readily available. The database among others contains information on all separate opinions, including the information which judges attached the separate opinion together. The main drawback of the CCC institutional setup is the inability to measure the role of political preferences of judges due to the lack of information on judges' votes [@martinDynamicIdealPoint2002; @hanrettyDissentIberiaIdeal2012] and the lack of variance in the nomination process.^[As explained bellow, practically all judges of one term have been nominated and appointed by the same President and Senate.] 

This article aims to fill this gap by conducting an empirical study of dissenting opinions at the CCC. We explore the factors that influence a judge's decision to dissent stemming from their strategic considerations and stemming from the Wittig's identification-disagreement model. We find that in line with the identification-disagreement model, the disagreement potential of any given case is positively correlated to the likelihood of a separate opinion. In contrast to that, we do not find evidence that the adherence to the norm of consensus varies across professional background of the judges. We also find that the workload a judge has is correlated with their likelihood of attaching a separate opinion. Lastly, we do not find evidence that the dissenting behavior varies depending on the time elapsed between the start of judges' terms and the date of a decision.

Our article proceeds as follows. In [section 3](#theory), we discuss the strategical and the identification-disagreement accounts of dissenting behavior. Therein, we draw hypotheses based on the general theory applied to the specific institutional setup of the CCC. In [section 3](#primer), we introduce the CCC. We discuss its institutional setup, its procedures and, most importantly, the rules concerning separate opinions.  In [section 4](#method), we discuss our method, our model, and the operationalization of variables.  In [section 5](#empirics), we empirically test the hypotheses and we discuss the results. [Section 6](#conclusion) concludes. 

# Theories of Dissenting Behavior {#theory}
In the following section, we describe the relevant theory for our paper. In [section 2.1](#accounts), we concisely introduce the general accounts of judicial decision-making. In [section 2.2](#dissenting-behavior), we switch focus to judicial decision-making in the context of separate opinions. More specifically, we discuss how strategic considerations impact dissenting behavior of judges and we touch upon the identification-disagreement model. Lastly, in [section 2.3](#identification-disagreement), we delve into Wittig's identification-disagreement model. We discuss its two main components, the degree of adherence to the norm of consensus as well as the disagreement potential.

## Overview of Accounts of Judicial-Decision Making {#accounts}
A separate opinion is generated in the process of judicial decision-making. The accounts of the nature of judicial decision-making have evolved over time. At first, judges were perceived as deciding simply by means of law. Although the originalists of today would still subscribe to such a view, over time, the perception of judges has changed. The attitudinal accounts posited that judges are policy oriented. In other words, judges follow their own ideas and preferences when deciding cases. A lot of research has been conducted on whether, how and to what extent do judges indeed seek to advance the policies they desire [@berdejoElectoralCyclesUS2017; @clarkLocatingSupremeCourt2010; @dworkinPoliticalJudgesRule1980; @kastellecEmpiricallyEvaluatingCountermajoritarian2016; @moyerJudicialInnovationSexual2012; @gschwendAreJudgesPolitical2016].

However, as of recently, the perspective on judges has shifted. Judges are now allegedly strategic and rational actors. One of the early pioneers of this approach @posnerWhatJudgesJustices1993 presents a simple model of judicial utility as function mainly of income, leisure and judicial voting. Further research followed the Posner mode and presented alternative models of judicial utility [based on economic psychology @foxallWhatJudgesMaximize2004]. Replacing the policy oriented approaches, which hold judges to pursue political policy oriented goals, researchers now focus more on their self-interest in terms of career progression, higher income, more leisure, or lesser workload [@epsteinStrategicRevolutionJudicial2000]. They argue that judges as "(1) social actors make choices in order to achieve certain goals, (2) social actors act strategically in the sense that their choices depend on their expectations about the choices of other actors, and (3) these choices are structured by the institutional setting in which they are made." [@epsteinStrategicRevolutionJudicial2000, p. 626] We now zero in on the theories of dissenting behavior as a subset of judicial decision-making.

## Accounts of Dissenting Behavior {#dissenting-behavior}
In their empirical study on dissenting behavior on the Supreme Court of the USA ("SCOTUS"), the proponents of the strategic account of judicial decision-making @epsteinWhyWhenJudges2011 base their theory of dissents on the strategic-economic framework of self-interested strategically motivated judges. In general, "a potential dissenter balances the costs and benefits of issuing a dissenting opinion." [@garoupaJudicialDissentCollegial2022; @garoupaDisagreeingPrivateDissenting2022]. @epsteinWhyWhenJudges2011 posit that judges have "leisure preferences, or, equivalently, effort aversion, which they trade off against their desire to have a good reputation and to express their legal and policy beliefs and preferences (and by doing so perhaps influence law and policy) by their vote, and by the judicial opinion explaining their vote, in the cases they hear." Therefore, a dissenting opinion comes at costs (trade offs) and comes with potential benefits that the judges strategically weight against each other.

The utility of a dissenting opinion are the potential to undermine the majority opinion when the dissent is influential and the enhanced reputation that the judge enjoys. The dissenting opinion may be cited in the future by other judges or publicly analysed by legal scholars. @epsteinWhyWhenJudges2011 also argue that the judges strategically take into account collegiality costs. The collegiality costs are lower at courts that sit in larger panels, whereas they are bigger at courts that decide in smaller panels as the judges have to spent time in a smaller circle of their colleagues. Moreover, they predict that the judges may reap benefits of averting a dissent whenever they face a high workload. In that, they free up their hand to take care of more pressing work. @epsteinWhyWhenJudges2011 then empirically verify to what extent does their theory hold. They find that the US judges indeed take into account the reputation utility, the collegiality costs as well as their workload.

In the European context, @wittigOccurrenceSeparateOpinions2016 in her dissertation thesis on separate opinions at the Federal Constitutional Court of Germany ("GFCC") summarizes the potential motivations for judges to attach a separate opinion and, thus, to acquire additional costs: (1) potential of impacting future caselaw, (2) moral obligation to distance oneself from a decision that contradicts her values, (3) to convey certain image about oneself. These motivations also largely rely on the self-perceived stance towards separate opinions in general. The proponents of separate opinions view dissenting positively based on the separate opinions  being able to enrich the legal debate, being a sign of judicial independence, increasing the legitimacy of any given decision for it makes the decision more accurate of the real discussion behind it. The opponents of separate opinions mainly argue that showing the inability to speak in one voice undermines a court's legitimacy or the reputation of the dissenting judge. Such a view is perfectly in line with the quote from the introduction. Moreover, judges seeking the appreciation from the general public or legal community may act in their personal interests instead of in the court's interests. Lastly, separate opinions come at collegiality costs and may harm the mutual relationships of judges.

In her theory of dissenting behavior, Wittig makes a sharp cut from the accounts coming mainly from the US, more specifically from the research on SCOTUS, and comes up with a model of separate opinions better suited for the civil law context of the CCC, the identification-disagreement model. Wittig argues that the traditional all have limited explanatory power as such and also do not fit within the civil law context, as judges therein are deciding in a different context, bound by different procedural rules, and, thus, given differing, sometimes broader, sometimes more limited, avenues to give way to their policy preferences or strategic considerations. We now discuss the identification-disagreement model in more detail.

## The Identification-Disagreement Model of Dissenting Behavior {#identification-disagreement}
Wittig introduces a non-formal model of separate opinions, the identification-disagreement model. Wittig amalgamates all the previously introduced potential motivations of judges for writing separate opinions into one cohesive and comprehensive model. The model is made up of two dimensions. The first dimensions of the model covers the disagreement level. The second dimension concerns the judges' stance and degree of self-identification of their role as a judge, Wittig terms this as a *norm of consensus*. Separate opinions are then "a function of a judge's identification with the norm of consensus and the level of disagreement of judges [@wittigOccurrenceSeparateOpinions2016, p. 74-75]. We base our theory on the identification-disagreement model and we use it to generate hypotheses for the CCC.

Following the identification-disagreement model, the probability of a separate opinion depends on judges' adherence to the norm of consensus and the level of disagreement. Therefore, the first two hypotheses are as follows:

**H~1~:** *The probability of observing a separate opinion is higher for judges with low norm-identification than for judges with high norm-identification.*

**H~2~:** *The probability of observing a separate opinion is higher for cases with a higher level of disagreement potential than in cases with a lower level of disagreement potential.*

We now delve into the two dimensions deeper. We start with the norm of consensus and then we move on to the disagreement.

### The norm of consensus
@calderiaTimeConsensualNorms1998, p. 876-877 define a norm as "a long-run equilibrium outcome, which underpins the interaction between individuals and reflects common understandings as to what is acceptable behavior in given circumstances." The norm of consensus in turn defines the level of dissent that is acceptable at any given court [@narayanConsensualNormHigh2005; @wittigOccurrenceSeparateOpinions2016, p. 75.]. Wittig's argument is two-fold. First, in civil law traditions unlike its US counterpart, the prevailing notion of the norm of consensus is that a court should not display disagreement. Second, the extent of adherence to the norm varies among judges,^[We conducted interviews with the judges of the third term of the CCC. Practically all of them more or less directly confirmed that they share the view that judges should not display dissent at a civil law court to a very varying degree.] depending on how they weight the costs and benefits they receive from following it [@wittigOccurrenceSeparateOpinions2016, p. 75.].

A dissonance between a proposed outcome for a case and any given judge's preferences are eventually bound to happen. In such a case, the judge can either express their sincere preferences by writing a separate opinion or they can adapt their behavior according to the norm of consensus and suppress the expression of her preferences. The second route has also been termed *dissent aversion* and theoretically fleshed out by @epsteinWhyWhenJudges2011. The decision of judge faced with such a conflict whether to attach a separate opinion or whether to avert their dissent is then a function of multiple potential utilities.

Wittig draws up three types of utility that dictate various levels of the adherence to the norm of consensus. Firstly, the intrinsic utility is maximized whenever a judge behaves in accordance with their true values and opinions, setting aside their strategic or political considerations. Secondly, expressive utility is harnessed when one displays individuality and counters the notion of conformism. Thirdly, the reputational utility arises when one adjusts their publicly displayed preferences to the expectations of others. Wittig argues that maximizing the former two forms of utilities in a situation of disagreement leads to separate opinions, whereas maximizing the reputational utility in such a situation gives way to the norm of consensus, as the judge would otherwise jeopardize the court's legitimacy as well as their reputation for not adhering to commonly accepted norms [@wittigOccurrenceSeparateOpinions2016, p. 76]. 

To some extent, we argue that even the third utility, the reputational utility, may lead to separate opinions insofar the individual reputation of a judge can in any way be linked to their non-conformity with the majority. An example that springs into mind is the late judge Scalia, whose individual reputation among conservative circles would've been likely more jeopardized by siding with the liberal majority rather than with not adhering to the norm of consensus [@scaliaDissents1998]. The decision to dissent or to avert a dissent then is a result of an weighting between costs and benefits of these three types of utilities a judge derives from adhering or not adhering to the norm of consensus.

How to capture the norm-identification of judges? The key dimension across which the extent of norm-identification may be theoretically expected to vary across judges is the professional background of the judges. The judges socialized within the judiciary are more likely to adhere to its values than those coming outside of it. In other words, CCC judges being appointed from the ranks of the judiciary are expected to adhere to the norm of consensus more than the rest. That applies especially strongly within the continental legal system, in which voicing a disagreement is usually prohibited within the ordinary judiciary. In Czechia, apart from the CCC and the Supreme Administrative Court, no separate opinions are even legally allowed. Therefore, the CCC judges coming out of the judiciary are socialized in an environment, in which voicing a dissent is not even allowed by law. On the other hand, lawyers or judges recruited from the ranks of the politicians are not socialized from within the judiciary and are used to defending their own opinion, or at least they are not required to hide their disagreement. However, we also theoretically expect that the difference between the professions should wane over time as the non-judicial judges get accustomed to the role of a CCC judge. Our hypothesis 3 suggests: 

**H~3~:** *The closer the date of the decision to the date at which the judge entered, the larger the difference between the probability of observing a separate opinion of judges coming from judiciary and of judges coming outside judiciary.*

Research into the effect of professional background of judges is no novelty. While evidence in the aforecited studies concludes in the favor of presence of such an effect, in the European context, @garoupaSpurredLegalTradition2020 in their comparative study of Slovenia and Croatia found no evidence for the influence of judges' background, such as their former career or education. The issue with the study is that its main research goal was to investigate the effect of political fractionalization on the dissenting behavior of judges based on the principal-agent account of judiciary. To achieve their research goal, they narrowed the selection of cases to only those, which included a political actor as a petitioner [@garoupaSpurredLegalTradition2020, p. 5], and later "controlled for" as many variables possible without elucidating as to why they are relevant for the relationship between political fractionalization and the dissenting behavior. Both of research choices are inductive to selection bias that accentuates rather than attenuates the main effect of interest in the former case. It is not surprise that in a model based on selection of only politically salient cases, the influence of certain aspects of judges' background appears weak in comparison to political influences.

### Disagreement on the bench
A disagreement on a bench arises when the opinions on the matter diverge during a discussion and a judge has a reason to object the majority view. For example, case characteristics play an important role. Cases with more value-laden or controversial topics may give raise to more disagreement, similarly highly complex cases leave more space for disagreement. The sources of disagreement are seemingly manifold. To elucidate them theoretically, we rely on the literature on case-space model.

The case-space model is a theoretical model developed in an attempt to model the idiosyncrasies of court decision-making, i.e. that a court is a body resolving disputes, cases [@landaDisagreementsCollegialCourts2007; @laxNewJudicialPolitics2011]. It differs from political science policy-space models in that that it incorporates the role of law. In the case-space model, the way to represent a case is by locating it in a n-dimensional space of of possible cases, the case-space. A case then denotes a legally relevant event that has occurred out of many that could have occurred. Put more simply, a case is a bundle of legally relevant facts. A legal rule then is a cut, a cut point when the case-space is one-dimensional, that divides the case-space. An individual disposition of a case is then the judgment of an individual judge of the case depending on their preference over legal rules (the cut point) and the location of the case in the case-space. In other words, if the case at hand falls to one side of the judge's cut point, then they vote for certain outcome, if it falls on the other side, they vote against it.

@landaDisagreementsCollegialCourts2007 draw from the case-space model multiple theoretical sources of disagreement. The first and clear source of disagreement among judges is that about facts. Different judges may place the facts of the case in the case-space differently. There other sources of disagreement: which dimensions should be relevant under any given legal rule to determine the disposition of a case, disagreement about "thresholds" within dimensions, and a couple of more sources of disagreement, which all can be summarized as a disagreement about the legal rule.

To this end, we simplify the model into two similar characteristics that concern the facts and the legal rule: case complexity and case controversy. Case complexity in our understanding refers to the number of legal issues that a case has touched upon. In line @landaDisagreementsCollegialCourts2007 we suppose that the more legal rules there is in play in any given case, the more room for disagreement about rules. We are also able to distinguish between different types of legal issues. Case controversy refers to the facts. There are subjects matters that we believe are more prone to disagreement. In sum, the effect of disagreement generated the H~2~ and we discuss details in the [section](#operationalization) on operationalization.

Within the CCC, we can observe a special example of circumstances giving rise to higher level of disagreement. Research on judicial coalitions at the CCC has revealed that its third term between 2013-2023 is rather polarized and that there are two voting blocs of judges that clash against each other [@chmelCoOvlivnujeUstavni2021; @smekalMimopravniVlivyNa2021; @vartazaryanSitOvaAnalyza2022].^[The Smekal et al. book goes so far to coin the first coalition as a more left-leaning and the second as a more right-leaning, whereas we are not convinced by this label.] The articles rely primarily on network analysis of the dissenting opinions in the plenary proceedings and make inferential conclusions based on a rather superficial descriptive analysis. We hypothesize that should the relationship from the plenary sessions indeed exist, they should also carry over to the 3-member panel hearings. If this shows to be true, it would provide further evidence to the two coalition theory of the CCC [@chmelCoOvlivnujeUstavni2021; @vartazaryanSitOvaAnalyza2022; @smekalMimopravniVlivyNa2021] as well as for the Wittig's identification-disagreement model. Consistent with our theoretical part, we believe that such an situation is theoretically a special case of circumstances with higher level of disagreement. Therefore, our next hypothesis suggests

**H~4~:** *Having a 3-member panel composed of members of both judicial coalitions increases judges' likelihood of a dissent.*

Concerning the state of the art research on the effects of disagreement on dissent, @wittigOccurrenceSeparateOpinions2016 found a positive effect of the legal complexity on the dissenting behavior of judges, similarly @brickerBreakingPrincipleSecrecy2017 found a similar effect in a study on the German, Polish, Latvian, and Slovenian constitutional courts.

### Strategic considerations
While we base our hypotheses mainly on Wittig's identification-disagreement model, we believe that strategic considerations as discussed by @epsteinWhyWhenJudges2011 can also come into play, albeit in a different institutional and regulatory setting than on SCOTUS. For example, workload varies across judge rapporteurs in the CCC context rather than across federal courts as in the US context. Therefore, we also base our one of our hypotheses on the strategic account of judicial decision-making. Rather than testing all potential sources of strategic considerations, we zero in on two specific sources of thereof, namely the role of collegiality costs and workload.

@epsteinWhyWhenJudges2011 address the issue of collegiality costs arising for a dissenting judge: "The effort involved in these revisions, and the resentment at criticism by the dissenting judge, may impose a collegiality cost on the dissenting judge by making it more difficult for him to persuade judges to join his majority opinions in future cases." Based on this theory, they predict and indeed empirically confirm that "dissents will be less frequent in circuits that have fewer judges because any two of its judges will sit together more frequently and thus have a greater incentive to invest in collegiality." Put simply, the researchers compare the dissent rates between courts with differing number of judges.

We test whether judges that are at the start of their term, and thus are aware that they will "sit together more frequently" invest in collegiality by averting separate opinions and whether when their term draws to an end, they give way to their disagreement. This presumes that the outlook of sharing the 10 year term with your colleagues at the beginning of judges' terms increases the collegiality costs of dissenting, whereas at the end of their terms, the collegiality costs decrease with the end of the shared term looming on the horizon. We draw our fourth hypothesis:

**H~5~:** *The closer the date of the decision to the date at which the judge entered the office, the lower probability of a separate opinion, whereas the closer the date of the decision to the date at which the judge left the office, the higher the probability of a separate opinion.*

The importance of leisure and workload on judicial decision-making has already caught attention of the empirically oriented legal scholarship. @clarkEstimatingEffectLeisure2018 have in a quasi-experimental difference-in-differences research design found that judges have preferences regarding their leisure, which then impacts their performance. They found that whenever a team of a US Federal Court judge's alma mater plays a basketball game, the time to process a decision increases for that respective judge and the quality of their work decreases. @brekkeThatOrderHow2023 found that the CJEU judges take into account their workload when issuing orders. @engelMannaHeavenJudges2020 found in a naturally occuring experiment in Israel that reduction in judges' affects the way they decide. In their paper on dissenting behavior of the SCOTUS judges, @epsteinWhyWhenJudges2011 found among others that judges strategically decide to write fewer dissenting opinions whenever the workload on their court increases. The evidence although overwhelming hasn't been one sided though. @songerExplainingDissentSupreme2011 found no relation between workload and dissent rate at the Supreme Court of Canada. 

According to @epsteinWhyWhenJudges2011: "[t]he economic theory of judicial behavior predicts that a decline in the judicial workload would lower the opportunity cost of dissenting [...]." Using the language of the identification-disagreement model, we believe leisure to be an example of the individual utility a judge may consider. Contrary to Wittig, we believe individual utility may also pull the other way: against a separate opinion. Therefore, our hypothesis 5 suggests:

**H~6~:** *The higher the workload of a judge, the lower the probability of them attaching a separate opinion.*

# A brief primer on the CCC {#primer}
We now introduce the CCC, its institutional and procedural background, its powers as well as its composition. The CCC consists of fifteen judges^[I will proceed with calling the CCC judges as "judges" in the SCOTUS fashion [@boatrightHistoryMeaningUse2018] as they technically sit on the highest court in Czechia.], out of which one is the president of the CCC, two are vice presidents and twelve associate judges [following the terminology of @kosarConstitutionalCourtCzechia2020]. These fifteen judges are appointed by the president of the Czech republic upon approval of the Senate, the upper chamber of the Czech two-chamber Parliament. The judges enjoy 10 years terms with the possibility of re-election; there is no limit on the times a judge can be re-elected. The three CCC functionaries are unilaterally appointed by the Czech president.

The appointment procedure is similar to how the SCOTUS judges are appointed as the procedure lays in the hands of the president of the republic and the upper chamber. The minimal requirement for a CCC nominee are 40 years of age, a clean criminal record, a finished legal education and experience in the legal field. Other than that, the nomination is left to the consideration of the President of the Republic. After a nomination, the nominee is firstly interviewed in the constitutional law committee of the Senate, which produces an unbinding recommendation for the plenary Senate hearing. The final binding decision is then made by simple majority of the Senate plenary hearing. This procedure has lead to a situation, in which there is very little variance as to the nominating background of the judges. First, there is no nominating political party akin to the US context or the Spanish context [@hanrettyDissentIberiaIdeal2012]. Second, because the court was established in 1993 and filled within roughly a year of its establishment and because the term of the Czech president is 5 years and all the 3 Presidents, who'd finished their term at the time of writing this article, have been elected twice (for ten years it total), each president has had the chance to appoint all the fifteen members of "their" CCC. Therefore, the first term of the CCC has been termed the Václav Havel, the second the Václav Klaus and the third Miloš Zeman terms of the CCC.

Regarding the competences, the CCC is a typical Kelsenian court inspired mainly by the German Federal Constitutional Court. The CCC enjoys the power of abstract constitutional review, including constitutional amendments. The abstract review procedure is initiated by political actors (for example MPs) and usually concerns political issues. Moreover, an ordinary court can initiate a concrete review procedure, if that court reaches the conclusion that a legal norm upon which its decision depends is not compatible with the constitution. Individuals can also lodge constitutional complaints before the CCC. Lastly, the CCC can also resolve separation-of-powers disputes, it can *ex ante* review international treaties, decide on impeachment of the president of the republic, and it has additional ancillary powers [for a complete overview, see @kosarConstitutionalCourtCzechia2020].

The CCC is an example of a collegial court. From the perspective of the inner organization, the CCC can decide in four bodies: (1) individual judges in the role of judge rapporteur, (2) 3-member chambers (*senáty*), (3) the plenum (*plénum*), and (4) special disciplinary chamber. The 3-member chambers and the plenum play a crucial role. The plenum is composed of all judges, whereas the four 3-member chambers are composed of the associate judges. Neither the president of the CCC or her vice-presidents are permanents members of the 3-member chambers. Until 2016, the composition of the chambers was static. However, in 2016, a system of regular 2-yearly rotations was introduced, wherein the president of the chamber rotates to a different every 2 years. I am of the view that such a institutional change opens up potential for quasi-experimental research similar to the @gschwendAreJudgesPolitical2016 study utilizing judge absences within the 3-member chambers of the German federal constitutional court. In general, the plenum is responsible for the abstract review, whereas the 3-member chambers are responsible for the individual constitutional complaints.

In the chamber proceedings, decisions on admissibility must be unanimous, whereas decisions on merits need not be, therefore, a simple majority of two votes is necessary to pass a decision on merits. In the plenum, the general voting quorum is a simple majority and the plenum is quorate when there are ten judges present. The abstract review is one of the exceptions that sets the quorum higher, more specifically to 9 votes.

A judge rapporteur plays a crucial role [@chmelZpravodajoveSenatyVliv2017; @horenovskyProcessMakingConstitutional2015 study the large influence of the judge rapporteurs at the CCC]. Each case of the CCC gets assigned to a judge rapporteur. The assignment is regulated by a case allocation plan.^[The original term is *rozvrh práce*, which is usually translated as a *work schedule*, however, I borrow the term *case allocation plan* from @hamannGermanFederalCourts2019, p. 673.] They are tasked with drafting the opinion, about which the body then votes. The president of the CCC (in plenary cases) or the president of the chamber (in chamber cases) may re-assign a case to a different judge rapporteur if the draft opinion by the original judge rapporteur did not receive a majority of votes. Unfortunately, the CCC does not keep track of these reassignments.

The act on the CCC allows for separate opinions. They can take two forms: dissenting or concurring opinions. Each judge has the right to author a separate opinion, which then gets published with the CCC decision. It follows that not every anti-majority vote implies a separate opinion, it is up to the judges to decide whether they want to attach a separate opinion with their vote. Vice-versa, not every separate opinion implies an anti-majority vote, as the judges can attach a concurring opinion. In contrast to dissenting opinion, when a judge attaches a concurring opinion, they voted with the majority but disagree with its argumentation.^[Which makes it difficult to, for example, conduct the same point-estimation with data on dissenting behavior of judges as @hanrettyDissentIberiaIdeal2012 has done on the Portugese and Spanish Constitutional Courts.] Unfortunately, it is difficult to conduct research on the dissent aversion because the voting is kept secret.

The room for the dissenting judge and the majority to address each other differs between the two bodies. Based on our qualitative research, there is less back and forth interplay between the judges, more akin to the SCOTUS context, and most of the communication is handled remotely in the panel proceedings, whereas the plenum meets regularly to discuss the cases in person. Despite that, procedurally speaking, the process of generating separate opinions is the same. In both cases, the rapporteurs are informed about the outcome of the vote, which is filed in the voting record. The separate opinion is then sent to the judge rapporteur before the decision is announced, as it cannot be added until after the announcement. It is important to note that judges have the possibility, not the obligation, to dissent. In other words, there is room for judges to give way to strategic considerations.

# Method {#method}
We now move onto the empirical part of our paper. First, in [section 4.1](#data-description) we describe the data we collected. Second, in [section 4.2](#operationalization) we explain how we operationalized the variables included in our hypotheses. Last, in [section 4.3](#identification) we explain our identification strategy.

## Data description {#data-description}
The data is based on the CCC dataset [@paulikCzechConstitutionalCourt2024], which contains all decisions published by the CCC since its foundation, complete text corpus as well as plenty of metadata. The whole dataset is available at [Zenodo](https://zenodo.org/records/10591110). We narrow our cases to all plenum decisions and to all 3-member chamber decisions on merits up until the end of 2022. The admissibility decisions of the 3-member chambers must be made unanimously, concurring decisions therein are a rarity.^[On top of that out of the 39 separate opinions in admissibility 3-chamber decisions, 25 of that are a copypasta from judge Jan Filip and 6 are a copypasta from judge Josef Fiala in alike cases. Thus, there is only a few left for a meaningful analysis. The class imbalance of the remaining ~10 decisions would be too large against all chamber decisions on admissibility] Fig. \@ref(fig:separate-opinion-ratio) confirms our intuition about the lack of variance among the admissibility decisions.

```{r separate-opinion-ratio, fig.cap = "Percentage of Decisions with at least one Separate Opinion"}
separate_opinion_ratio
```

We skip the first decade of the CCC as the data on it are rather incomplete and inconsistent: For example, very few decisions contain the information on the composition in the text and many do not contain the name of the dissenting judge at all. We also limit our analysis until the end of 2022 as the CCC entered its 4th term in 2023 and started to undergo a complete personal change. 

The final dataset for analysis contains `r toString(length(unique(data$doc_id)))` decisions of the CCC. Table \@ref(tab:overalltable). reveals that out of them, `r overall_table[overall_table$formation == "Chamber", "ratio_total"]` are decisions by the 3-member chambers on merits, around `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "merits", "ratio_total"]` are plenum decisions on merits and the remaining `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "admissibility", "ratio_total"]` are plenum decisions on admissibility. At least one separate opinion contain `r overall_table[overall_table$formation == "Chamber", "ratio_dissent"]` decisions out of the 3-member decisions, `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "admissibility", "ratio_dissent"]` decisions of the plenum decisionson admissibility, and `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "merits", "ratio_dissent"]` decisions of the plenum decisions on merits.

```{r overalltable}
overall_table %>%
  kable(
    caption = "Summary statistics for the whole dataset", 
    col.names = c("Formation", "Grounds", "Count", "%", "% with at least One Separate Opinion")
  ) %>%
  row_spec(0,bold=TRUE) %>%
  kable_styling()
```


## Operationalization {#operationalization}
We now discuss how we operationalized the variables included in our hypotheses. We start with the dependent variables, then we move on to the explanatory variables and finish with control variables.

### Dependent variable: a separate opinion
We conceptualize our dependent variable as the information whether a judge attached a separate opinion to a decision or not.^[Unlike Wittig we do not call our dependent variable a judge's vote, as that refers to a slightly different thing within the CCC context. A judge may vote against the majority opinion but since they are not mandated to write a separate opinion, these do not necessarily overlap. Similarly, a judge may vote for a disposition of a case and still attach a concurrence separate opinion.] Our dependent variable *separate opinion* is thus a dummy variable that has two categories: either a judge did attach a separate opinion (1) or she did not (0) in any given case .

We do not distinguish between a concurrence and a full dissent. The reason is two-fold: theoretical and practical. Theoretically, the difference between the two lays only in the disposition of the case. The judges may equally disagree on the interpretation of legal rules, thus, in the case-space model terms [@landaDisagreementsCollegialCourts2007; @laxNewJudicialPolitics2011], the judge cut points in any given case differ even when a judge attaches "only" concurrence. The difference is that in the cases containing concurrence, the case facts may have completely accidentally fallen on the same side both of the concurring judge as well as the majority, whereas in the cases containing a dissenting opinion, the case fell in between the cut points. We do not consider this phenomenon as theoretically interesting. Practically, it was impossible to distinguish between the two categories as the texts of especially the older decisions simply do not contain the information whether a separate opinion was a concurring or dissenting opinion.

### Explanatory variables
#### Disagreement potential {-}
From the theoretical perspective it may be expected that the potential for disagreement varies across cases. In some cases, the disagreement potential is higher and, thus, the probability of a separate opinion is higher than in the cases with lower potential for disagreement. More specifically, based on our theory we expect the potential for disagreement to be captured by two characteristics of any given case: (1) its legal complexity and (2) its controversy or in terms of @epsteinMeasuringIssueSalience2000 issue salience.

##### Complexity {-}
As @corleyPuzzleUnanimityConsensus2013 argue and empirically measure, complexity of a case leads to less certainty and more ambiguity for the judges, which leads to a higher probability of disagreement. The authors define legal complexity as the number of legal issues a case has to address. True to the @corleyPuzzleUnanimityConsensus2013 study, our operationalization of complexity relies on the assumption that the more legal issues there are in any given case, the higher the number of references to other laws and caselaw in the text of the corresponding decision. What's more, we are able to distinguish between legal issues at the constitutional and ordinary level.

The variable *concerned acts* captures the number of concerned ordinary legal acts on the legal-act level, the variable *concerned constitutional acts* captures the number of articles of the constitutional legal acts (mostly the Czech Constitution and the Charter of Fundamental Rights and Freedoms) and variable *caselaw* captures the number of citations to its own caselaw. The information on the first two variables is based on the metadata provided by the CCC, the last is extracted via a regular expressions search of the text of the decisions.

```{r}
corr_acts %>%
  as_tibble() %>%
  mutate(
    term = c("1. # of Ordinary Acts", "2. # of Constitutional Articles", "3. # of CCC References")
    ) %>%
  kable(
    caption = "Correlations between Independent Variables of Case Complexity",
    col.names = c("Variable", "1", "2", "3")
  ) %>%
  row_spec(0,bold=TRUE) %>%
  kable_styling() %>%
  footnote(
    general = "Correlation was calculated using the Spearman Correlation Coefficient.",
    general_title = "Note.",
    footnote_as_chunk = TRUE
    )
```

The first two variables we believe to be sufficiently different from each other: a typical right to fair proceedings case may concern only one constitutional article (the article 36 of the Charter) but many legal acts, whereas a typical separation of powers case concerning the Parliament concerns many constitutional articles but only few ordinary laws. On the other hand, the number of (concerned) constitutional acts and references to the CCC caselaw may be correlated. We therefore run diagnostics, which reveal that the citations to CCC caselaw and the number of references to constitutional acts are rather correlated. Because we believe the number of cited constitutional articles to better reflect the number of legal issues than the number of references to CCC case-law, in our final model we exclude the number of references variable.

##### Controversy/Issue salience {-}
In a similar vein, certain typically value-laden issues may generate more disagreement even if they raise only one or few legal questions. @epsteinMeasuringIssueSalience2000 distinguish between two types of issue salience. Retrospective issue salience is how a particular issue is viewed in hindsight, regardless of whether the actors at the time regarded it so. Contemporaneous salience than captures the stance of the relevant actors, the CCC judges in our cases, on the issue. While we are interested in the latter, it is not obvious as to how exactly can such an information be captured. @epsteinMeasuringIssueSalience2000 reject all the typically employed proxies for issue salience such as the number of articles in law review that a case generated or the number of citations a case generated. Instead, they propose a novel unbiased transparent measure of issue salience: whether a SCOTUS decision made it to the front-page of the New York Times newspaper. Unfortunately, practically all of the measures are hard to transfer to a non-US context. We cannot exactly pinpoint a similar newspaper as New York Times in Czechia. Therefore, we need to rely on another measure of issue salience/case controversy.

In her thesis, @wittigOccurrenceSeparateOpinions2016 selected a number of subject matters she deemed as particularly controversial. In the CCC context, there are a number of issue that have been deemed as controversial by the Czech legal scholarship. Typically, the restitution cases or cases concerning fundamental human rights have been coined as rather controversial. The CCC database contains a table *subject_matter*, which contains the subject matter of any given case (a discrimination case, a separation of powers case and alike) and information on the area of constitutional law that the case pertains (the right to fair trial, the right of freedom of speech). We coded the subject matter as controversial when they concern any of the following subject matters: discrimination, expropriation, restitution, the property of church, sexual orientation, the protection of consumer, fundamental human rights, social and cultural rights, the right of property, the freedom of speech, and the separation between the church and state. Therefore, the variable *controversial* is a dummy variable, which takes the value of 1 when the given case falls in any of the previously mentioned categories or it takes the value 0 when it does not.

#### Norm-identification {-}
Secondly, our theory generated a second expectation: a relationship between adherence to the norm of consensus and the likelihood of a separate opinion. The adherence to the norm of consensus depends on self-identification and on self-perception of the judges: "being part of the system of the judiciary influences identification with the organization's goals and it shapes considerably the judges' perception of how to behave." [@wittigOccurrenceSeparateOpinions2016, p. 100]. Theoretically, we expect the adherence to the norm of consensus to depend on one's career choices. The actors socialized within the judiciary and its values are more likely to adhere to them than their peers who entered the CCC from the legal practice, from politics, or from anywhere outside the judiciary [@wittigOccurrenceSeparateOpinions2016, p. 84-87.].

```{r age-distribution, fig.cap = "Kernel density of judges' ages at the end of their terms."}
age_distribution
```

Wittig confirms the theoretical intuition empirically as well but she operationalizes the norm-identification as the judges' career choice after their term. The judges that chose to stay within judiciary or go back to being scholars are expected to more strongly identify with the norm of consensus, whereas the judges' that made different career choice are less likely to identify with the norm of consensus. Unfortunately, such a measure does not fit the CCC context. As the table \@ref(fig:age-distribution), CCC judges start their term usually at the end of their career and a considerable part of them ends their term in their 70s, well past the retirement age in Czechia. Given that the retirement age of judges 65 years of age, we expect majority of the CCC judges to simply retire after their terms.

```{r profession-distribution, fig.width=6, fig.cap = "The distribution of professions of the CCC judges across its 3 terms."}
professions
```

Therefore, instead of operationalizing the norm-identification as the profession after their term, we operationalize it as the last profession before they entered the CCC. Viewed this way, we presume that the last profession that a judge had held before entering an office had the highest impact on their self-identification as a judge. Overtime we expect the effect to wane off. 
@garoupaDisagreeingPrivateDissenting2022 have found evidence that the individual decision to attach a separate opinion of the judges on the Spanish Council of State depends on the judge's professional background. The effect does not go in the direction that we would theoretically expect: only former politicians seem to be less likely to attach a separate opinion, while former judges and lawyers seem to be more likely to attach a separate opinion.

Quite handily, the CCC dataset contains such an information on the last profession before the CCC judges took up their office. In the dataset, the variable *judge_profession* contains the information on the judges' previous career choices and can take up the values of judge, scholar, politician, or lawyer. The table  \@ref(fig:profession-distribution) reveals that the distribution of the professions has changed across time. For our purposes, we flatten the original variable into 2 levels. The variable *profession* is coded either as "within" if the judge previously served as a judge or "outside" if they came from the other three professions.

#### Time in office {-}
The effects of the CCC judges' profession should according to our theoretical expectations vary depending on the time the judges have spent in their office. The closer to them taking up the office, the effects of their preceding profession should be more pronounced. Similarly, according to the collegiality costs hypothesis, the dissenting behavior of judges should vary across time. Namely, judges are expected to dissent less likely at the beginning of their terms, as the collegiality costs are higher, and to dissent more likely at the end of their terms, as the collegiality costs of the disagreement are lower. We capture the *time in office* variables as the number of months a judge has left until their end of mandate at the date of the decision. That allows us to account both for the collegiality costs hypothesis as well as the difference between professions hypothesis.

#### Workload {-}
Similarly to the study of @brekkeThatOrderHow2023 on workload at the CJEU, we operationalize *workload* as the number of pending unfinished cases that any given judge has in the moment of any given decision as a judge rapporteur. We used the information on the composition of the bench in any given decision from the CCC database. We then calculated the number of unfinished cases each judge had at the time of any given decision as a judge rapporteur using the date of submission and of decision of a case. The mean number of pending cases of any judge as a judge rapporteur is `r descr_workload$numeric.mean[1]`, with quite a large standard deviation of `r descr_workload$numeric.sd[1]`. The highest number of pending cases in the dataset is `r descr_workload$numeric.p100[1]` in the case of the judge Radovan Suchánek, whose work ethic has already become a topic of scholarly discussion [@chmelCoOvlivnujeUstavni2021].

```{r workload}
descr_workload %>% 
  kable(
    caption = "Summary statistics for the workload variable", 
    col.names = c("", "Mean", "SD", "P0", "P25", "P50", "P75", "P100"),
    digits = 0
  )
```

<!-- One looming issue arises that needs to be addressed and may introduce bias. Not every unfinished pending case is the same. Different judges may cope with the increasing workload differently. Czech scholars have argued that deciding cases on admissibility, which are typically shorter and require a rather concise argumentation, may be such a coping mechanism of overburdened judges: it allows them to expedite decisions faster and with less effort. Thus, the same number of unfinished pending cases of a judge that chooses the strategy of offloading their workload burden into decisions on admissibility may not carry the same workload as with a judge that opts for decisions on merits more often. As a solution, we add this ratio in the name of *ratio_admissibility* into our model as a control variable. -->

#### Mixed coalition
Lastly, we include a dummy variable mixed coalitions, which takes up the value 1 when two conditions are met: (1) the decision was made in a 3-member chamber and (2) the composition of the chamber was made up of judges from both coalitions.

## Identification Strategy {#identification}
The hypotheses are tested by fitting a generalized linear model estimating the probability of a judge attaching a separate opinion with the dependent variable following a binomial distribution. 

``` {r model-selection}
model_selection %>%
  kable(caption = "Descriptive statistics for the formation clusters",
        col.names = c("Units", "Mean Observations per Unit"))
```

When identifying our model, we stood before a research choice that has been addressed differently by different researchers. As other studies on judicial decision-making relying on observational data, the data on judicial-decision making offers multiple potential clustering variables to include fixed effects on. In their study of effect of french language on the performance of the judges at the CJEU, @cheruvuHowInstitutionalConstraints2019 used fixed effects in their model the effects both on the individual judges as well as on the panel level. The reasoning behind fixed effects clustered on the formation were that it "[s]cholars argue the number of judges sitting on a chamber is a proxy for a cases’ salience as more important cases tend to be assigned to larger chambers (e.g. @kelemenPoliticalFoundationsJudicial2013; @larssonJudicialIndependencePolitical2016). Including chamber fixed-effects in my models addresses both heterogeneity in the collegial decision-making process of different chambers of judges and implicitly controls for the number of judges hearing a case." @clarkEstimatingEffectLeisure2018 in their paper studied the effect of leisure on judicial performance.^[Also measured as time for a judge to decide a case similarly to the @cheruvuHowInstitutionalConstraints2019 study.]. They included the the author fixed effects in their model.^[Which would correspond to the judge rapporteur in our case.] @brekkeCJEUDatabasePlatform2023 in a study on the usage of orders at the CJEU built a logistic regression model with subject matter fixed effects.

Because in our study, we are interested in the between variance of individual judges as well as the subject matters, it does not make sense to include fixed effects of either cluster. In contrast to that, we are not interested in the so-called panel effects, especially the difference between the 15-member plenum  and the 3-member panels could produce bias. The more controversial cases or more legally complex cases are typically assigned to the plenum and the occurrence of separate opinions is higher in the plenary decisions. Moreover, the behavior of the 3-member panels varies as well. It thus makes the most sense to include panel fixed effects as that would allow us to control for a potentially potent source of bias. In our model, the variable *formation* captures the information on the 3-member panels, whose composition is relative unstable especially after the introduction of system of rotations of judges in 2016, therefore the information is coded as the exact composition of any given 3-member panel, and the information whether the case was decided by the plenum, whose composition is relatively stable within the two terms.

Moreover, the included data spans two terms of the CCC. Its second decade roughly between 2003 and 2013 and its third decade between 2014 and 2023. The composition of the CCC has completely changed between the two terms and so has, we expect, changed the behavior of the court as such. Rather than including year fixed effects, we opt for the term fixed effects. If the majority of judges sitting in a panel came from the second term, the variable *term* is then coded as "2nd", and if the majority of judges sitting in a panel came from the third term, the variable takes up the value "3rd".

We also build a second model for the purpose of our 4th hypothesis. To this end, because we are interested in information on the panel level, our unit of observation is flattened to a decision level. Our dependent variable is the information whether a separate opinion occurred or not in that decision as in the 3-member chamber decisions, the decisions have to be either made unanimously or with a 2 votes majority. Our explanatory variable is the information whether the panel was composed fully of either coalition or whether the composition was mixed. Because case assignment^[The cases are assigned randomly to judges rapporteur based on the first letter of their surname.] as well as the assignment of judges to panels is as good as random, and thus there is little room for confounding, we build only a very simple logistic model with data filtered only to include 3-member panel decisions within the 3rd term of the CCC. The model in itself is more of a robustness check of the conclusions of Czech legal scholarship rather than a full fledged model built for causal inference.

# Results and discussion {#empirics}
The results summarised in  tab. \@ref(tab:model-summary) reveal quite an interesting trend at the CCC. We tested the model under different specifications by excluding different variables. The full model presented  the best fit. However, while excluding the disagreement potential variables produced a significant drop in AIC, excluding the norm-identification did not. Therefore, the norm-identification variables do not seem to carry too much explanatory power regarding the dissenting behavior of CCC judges. The confusion matrix in tab. \@ref(tab:prediction) reveals that the model does a reasonable job predicting the original data.

```{r model-summary, warning = FALSE, echo = FALSE}
modelsummary::modelsummary(list("Full Model" = model_me_formation,
                                "Disagreement" = model_me_disagreement,
                                "Norm" = model_me_identification),
                           output = "kableExtra",
                           estimate = "{estimate}{stars}",
                           statistic = "({std.error})",
                           stars = TRUE,
                           title = "Results from the Logit Model",
                           coef_rename = c("Intercept", "N Concerned Acts", "N Concerned Constitutional Acts", "Admissibility", "From Within Judiciary", "Time in Office", "Controversial", "Workload", "From Within Judiciary X Time in Office", "SD (Intercept Formation)", "SD (Intercept Term)")) %>%
  kable_styling(latex_options = "HOLD_position")
```

Moving on to the interpretation of the results, our results do not support the Hypothesis 1. Unlike at the GFCC, we do not find evidence for the norm-consensus operating according to our theoretical expectations. Firstly, adding the interaction term between the professional background and the time in office does not improve the model fit. It does not carry much explanatory power as an ANOVA test fails to reject the null hypothesis of no difference. Secondly, neither the separate variables of norm-identification nor the interaction term are near statistical significance. Fig. \@ref(fig:predicted-probabilities) plots the predicted probability conditional upon the professional background and the time in the office. The figure reveals that although the behavior of judges with differing professional background converges over time, the difference is insignificant. That does not however lead to us rejecting the identification-disagreement as a theoretical model. We believe it to be a clear theoretical model that can guide further empirical inquiries. It may just be the case that the proxy of the judges' profession may not be wholly accurate or that even if it was there may simply be currently no norm of consensus operating at the CCC. 

```{r prediction}
conf_matrix$table %>%
  as_tibble() %>%
  kable(caption = "A confusion matrix (in a long tabular format) of precitions produced by the GLM model. Where the prediction and the truth are the same, the model predicted the outcome correctly and vice-versa.",
        col.names = c("Prediction", "Truth", "N"))
```


In contrast to that, we find a clear support for Hypothesis 2: all the disagreement potential variables are positively and significantly correlated with the likelihood of a judge attaching a separate opinion. No matter the model specification, we always found a clear statistical, substantive and positive relationship between the probability of a judge attaching a separate opinion and the number of constitutional or ordinary acts concerned.

``` {r placebo-test}
placebo_models %>%
  mutate(term = paste0("Placebo", 1:10)) %>%
  select(-statistic) %>%
  kable(caption = "The result of 10 models fitted with different placebo sample from the subject matter variable.",
        col.names = c("", "Estimate", "SE", "p.value"),
        digits = 2) %>%
  kable_styling(latex_options = "hold_position")
```

The same applies to the controversial value-laden topics. Because we are aware of the potential issue with "arbitrarily" selecting a couple of potential subject matters, we decided to test the robustness of our *controversial* variable. To do so, we conducted a placebo test. We sampled 10 placebo samples of different subject matters (n = `r length(controversial_topics)`) and fitted the same final full mixed-effects model with the *controversial* variable being replaced by the placebo. After rerunning the placebo test multiple times with various samples of different seeds, we found that only sometimes would at most one placebo sample out of the 10 be statistically significant, the remaining 9 or 10 samples were not, tab. \@ref(tab:placebo-test) displays one of those runs.  While such a conclusion may not come as a complete surprise, as it is perfectly aligned with theoretical expectations, we provide a firm evidence for it.

```{r predicted-probabilities, fig.cap = "Predicted probability of a separate opinion occuring conditional on the previous profession of a judge and the time left in their term. Time in office refers to the months left in the office."}
marginal_probabilities +
  ggtitle(NULL)
```


In line with our hypothesis 4 based on the theoretical expectations generated by the Czech legal scholarship as well as the disagreement potential, we find that a separate opinion is more likely to occur in the 3-member chamber cases, whose composition is mixed by members of both voting blocs from the plenary proceedings. Tab. \@ref(tab:model-coalitions) reveals that the size of the effect is not negligible. While we are aware that our model is not thorough enough for any (causal) inference, it is rather a form of robustness check to the literature generated by the Czech legal scholarship that we'd summarized in the theoretical section.

```{r model-coalitions}
modelsummary::modelsummary(model_coalitions,
                           estimate = "{estimate}{stars}",
                           statistic = "({std.error})",
                           output = "kableExtra",
                           stars = TRUE,
                           title = "Results from the Logit Model on Coalitions",
                           coef_rename = c("Intercept", "Same Coalition")) %>%
  kable_styling(latex_options = "hold_position")
```


Our results also support for Hypothesis 5. Higher workload of a judge decreases their probability of attaching a separate opinion. It thus appears that the claims of @epsteinWhyWhenJudges2011 are not unwarranted and may carry over to other contexts. The results reveal that judges of the CCC give way to strategical considerations regarding their leisure. If they're overburdened with work, they reduce the load elsewhere, namely in the additional burden of writing separate opinions.

Lastly, we do not find evidence for the so-called freshman's effect. The importance CCC judges attach to collegiality costs of a separate opinion seem to remain consistent across time as the time in office of a judge does not carry explanatory power in our model.

# Conclusion {#conclusion}
We examined the factors influencing dissenting opinions by judges at the CCC. Our findings do not support the existence of a strong norm of consensus at the CCC, unlike the GFCC. However, the complexity of a case, measured by the number of legal acts involved, is positively correlated with the likelihood of a dissenting opinion. Similarly, cases concerning controversial topics are more likely to generate dissents.  A placebo test strengthens this finding by demonstrating that randomly chosen, non-controversial topics have minimal impact on dissenting behavior. The effect of voting blocs in the plenary proceedings seems to carry over to the 3-member panel proceedings. The results also reveal that judges make strategic considerations. When facing a high workload, judges are less likely to write separate opinions, suggesting they prioritize workload management. Interestingly, the time a judge has been in office does not seem to influence their likelihood of dissenting, indicating a consistent level of importance placed on collegiality.

\vspace{30pt}

# References {-}

<div id="refs"></div>