---
output: 
  stevetemplates::article2:
    keep_tex: false
    toc: true
    number_section: true
    fig_caption: yes
title: "Disagreement and dissent on a bench: a quantitative empirical analysis of the Czech Constitutional Court"
author:
- name: Štěpán Paulík
  affiliation: Humboldt Universität zu Berlin, stepan.paulik.1@hu-berlin.de
- name: Gor Vartazaryan
  affiliation: Charles University, gorike2000@gmail.com
anonymous: FALSE
removetitleabstract: FALSE
abstract: ""
thanks: "Replication files are available on the author's Github account (https://github.com/stepanpaulik/apex_courts_dataset/). **Current version**: `r format(Sys.time(), '%B %d, %Y')`"
keywords: "empirical legal research, courts, dissents, judicial behavior, political science, regression analysis"
date: "`r format(Sys.time(), '%B %d, %Y')`"
link-citations: yes
linkcolor: blue
geometry: margin=1in
fontsize: 11pt
doublespacing: TRUE
endnote: no
# pandocparas: TRUE
sansitup: FALSE
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{float}
bibliography: "`r rbbt::bbt_write_bib('bibliography.bib', overwrite = TRUE)`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(patchwork)
library(tidyverse); theme_set(theme_minimal())
library(ggplot2)
library(kableExtra)
library(modelsummary)
library(bookdown)

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, fig.width=4.2, fig.asp=0.8)

load("../data/descriptive_statistics.RData")
load("../data/model_results.RData")
```

# Introduction
Empirical legal research has been slowly but surely finding it's outside the predominant US context. Historically though most of the empirical studies have been conducted in the US, especially the Supreme Court, context [such as @boydUntanglingCausalEffects2010; @carrubbaWhoControlsContent2012; @epsteinWhyWhenJudges2011]. We now know that judgments are what judges had for a breakfast. Put less pompously, there are many theories and approaches for explanation of judicial decision-making behavior stemming mainly from the US context [@posnerHowJudgesThink2010]. What we do not know is the extent to which these theories and explanations carry over to other legal systems and context.

Although it has been traditionally espoused that there has been a divide between the empirically oriented US legal scholarship, stemming from a different perception of the role of courts and judges, and the rest of the world [@hamannGermanFederalCourts2019, p. 416]. Therein the judges empirically researched whether and to what extent they behave as for example political or strategic actors.[@carrubbaWhoControlsContent2012; @clarkLocatingSupremeCourt2010; @epsteinChoicesJusticesMake1997; @lauderdaleScalingPoliticallyMeaningful2014; @sunsteinAreJudgesPolitical2006; @cameronChapterWhatJudges2017; @clarkEstimatingEffectLeisure2018; @epsteinWhyWhenJudges2011; @epsteinStrategicRevolutionJudicial2000; @kornhauserModelingCollegialCourts1992; @kornhauserModelingCollegialCourts1992a; @posnerWhatJudgesJustices1993; @posnerHowJudgesThink2010; @rousseyOverburdenedJudges2018].

In contrast to, especially in European legal systems, such as the one at hand - Czechia, judges have been perceived as "proclaimers of law" and the law handed down by them [@hamannGermanFederalCourts2019, p. 417]. Such a view had hindered empirical legal research in Europe. At least so the story goes until recently. The interest in empirical legal studies has picked up in the last years across the whole continent, including studies on plethora of topics within Germany [@arnoldScalingCourtDecisions2023; @coupetteQuantitativeRechtswissenschaft2018; @engstEinflussParteinaheAuf2017; @wittigOccurrenceSeparateOpinions2016], Spain and Portugal [@hanrettyDissentIberiaIdeal2012], the UK [@hanrettyCourtSpecialistsJudicial2020] or the EU institutions [@bielenBacklogsLitigationRates2018; @brekkeThatOrderHow2023; @fjelstulHowChamberSystem2023; @fjelstulEvolutionEuropeanUnion2019; @fjelstulTimelyAdministrationJustice2022].

Such a view also goes against the possibility of a dissenting voice. A dissent shows lack of unity at the court. How that may be when a court is a proclaimer of law? Unsurprisingly thus, the possibility of attaching a separate opinion allowed mainly at the constitutional courts in Europe has already caught scholarly attention. In the US context, @epsteinWhyWhenJudges2011 have tested to what extent do US judges behave strategically when attaching separate opinions. In the European context, @kelemenJudicialDissentEuropean2017 offers a mainly theoretical comparative overview of the various regimes of dissenting behavior across European courts. Hanretty has made use of dissenting behavior of Spanish and Portugese judges to conduct a point estimation of location of the location of the judges [@hanrettyDissentIberiaIdeal2012] and to investigate what does a dissent reveal about the dimension across which the disagreement runs on Estonian Supreme Court [@hanrettyJudicialDisagreementNeed2015]. Most importantly, @wittigOccurrenceSeparateOpinions2016 has put forward and empiricall tested a theoretical model of dissenting behavior tailored at the European context that aims to overcome the shortcomings of the traditional accounts of judicial behavior stemming from the US context.

In our article, we pick up on this change and we set out to conduct an empirical research into the circumstances of disagreement on a court bench, we empirically research the circumstances of disagreement among the judges of the Czech Constitutional Court ("CCC") and we attempt to square the perception of judges as strategic rational-economic beings [@epsteinWhyWhenJudges2011; @epsteinStrategicRevolutionJudicial2000], the identification-disagreement model [@wittigOccurrenceSeparateOpinions2016], and our empirical findings.

We find that in line with the identificatgion-disagreement model, the disagremenet potential of any given case is positively correlated to the likelihood of a separate opinion. In contrast to that, we do not find evidence that the adherence to the norm of consensus varies across professional background of the justices. We also find that the workload a justice has is correlated with their likelihood of attaching a separate opinion. Lastly, we do not find evidence that the dissenting behavior varies depending on the time elapsed between the start of justices' terms and the date of a decision.

Our article proceeds as follows. In [section 2](#theory), we lay out theory of dissenting behavior. We mainly focus on the disagreement-identification model. In [section 3](#primer), we introduce the CCC. We discuss its institutional setup, its procedures and, most importantly, the rules concerning separate opinions. In [section 4](#hypotheses), we draw hypotheses based on the general theory applied to the specific institutional setup of the CCC. In [section 5](#empirics), we empirically test the hypotheses. We describe our data, we explain how we operationalized our variables, we describe our identification strategy, and lastly we discuss the results. [Section 6](#conclusion) concludes. 

# Theory {#theory}
In the following section, we describe the relevant theory for our paper. In [section 2.1](#accounts), we concisely introduce the general accounts of judicial decision-making. In [section 2.2](#dissenting-behavior), we switch focus to judicial decision-making in the context of separate opinions. More specifically, we discuss how strategic considerations impact dissenting behavior of judges and we touch upon the identification-disagreement model. Lastly, in [section 2.3](#identification-disagreement), we delve into Wittig's identification-disagreement model. We discuss its two main components, the degree of adherence to the norm of consensus as well as the disagreement potential.

## Overview of Accounts of Judicial-Decision Making {#accounts}
A separate opinion is generated in the process of judicial decision-making. The accounts of the nature of judicial decision-making have evolved over time. At first, judges were perceived as deciding simply by means of law. Although the originalists of today would still subscribe to such a view, over time, the perception of judges has changed. The attitudinal accounts posited that judges are policy oriented. In other words, judges follow their own ideas and preferences when deciding cases. A lot of research has been conducted on whether, how and to what extent do judges indeed seek to advance the policies they desire [@berdejoElectoralCyclesUS2017; @clarkLocatingSupremeCourt2010; @dworkinPoliticalJudgesRule1980; @kastellecEmpiricallyEvaluatingCountermajoritarian2016; @moyerJudicialInnovationSexual2012; @gschwendAreJudgesPolitical2016].

However, as of recently, the perspective on judges has shifted. Judges are now allegedly strategic and rational actors. One of the early pioneers of this approach @posnerWhatJudgesJustices1993 presents a simple model of judicial utility as function mainly of income, leisure and judicial voting. Further research followed the Posner mode and presented alternative models of judicial utility [based on economic psychology @foxallWhatJudgesMaximize2004]. Replacing the policy oriented approaches, which hold judges to pursue political policy oriented goals, researchers now focus more on their self-interest in terms of career progression, higher income, more leisure, or lesser workload [@epsteinStrategicRevolutionJudicial2000]. They argue that judges as "(1) social actors make choices in order to achieve certain goals, (2) social actors act strategically in the sense that their choices depend on their expectations about the choices of other actors, and (3) these choices are structured by the institutional setting in which they are made." [@epsteinStrategicRevolutionJudicial2000, p. 626] We now zero in on the theories of dissenting behavior as a subset of judicial decision-making.

According to standard individual rational dissent theory [@garoupaJudicialDissentCollegial2022; @garoupaDisagreeingPrivateDissenting2022], a potential dissenter balances the costs and benefits of issuing a dissenting opinion (the analysis also applies to concurring opinions, where costs and benefits are likely to be smaller in magnitude)

## Accounts of Dissenting Behavior {#dissenting-behavior}
In their empirical study on dissenting behavior on the Supreme Court of the USA ("SCOTUS"), the proponents of the strategic account of judicial decision-making @epsteinWhyWhenJudges2011 base their theory of dissents on the strategic-economic framework of self-interested strategically motivated judges. In general, "a potential dissenter balances the costs and benefits of issuing a dissenting opinion." [@garoupaJudicialDissentCollegial2022; @garoupaDisagreeingPrivateDissenting2022]. @epsteinWhyWhenJudges2011 posit that judges have "leisure preferences, or, equivalently, effort aversion, which they trade off against their desire to have a good reputation and to express their legal and policy beliefs and preferences (and by doing so perhaps influence law and policy) by their vote, and by the judicial opinion explaining their vote, in the cases they hear." Therefore, a dissenting opinion comes at costs (trade offs) and comes with potential benefits that the judges strategically weight against each other. 

The utility of a dissenting opinion are the potential to undermine the majority opinion when the dissent is influential and the enhanced reputation that the judge enjoys. The dissenting opinion may be cited in the future by other judges or publicly analysed by legal scholars. @epsteinWhyWhenJudges2011 also argue that the judges strategically take into account collegiality costs. The collegiality costs are lower at courts that sit in larger panels, whereas they are bigger at courts that decide in smaller panels as the justices have to spent time in a smaller circle of their colleagues. Moreover, they predict that the justices may reap benefits of averting a dissent whenever they face a high workload. In that, they free up their hand to take care of more pressing work. @epsteinWhyWhenJudges2011 then empirically verify to what extent does their theory hold. They find that the US judges indeed take into account the reputation utility, the collegiality costs as well as their workload.

In the European context, @wittigOccurrenceSeparateOpinions2016 in her dissertation thesis on separate opinions at the Federal Constitutional Court of Germany ("GFCC") summarizes the potential motivations for judges to attach a separate opinion and, thus, to acquire additional costs: (1) potential of impacting future caselaw, (2) moral obligation to distance oneself from a decision that contradicts her values, (3) to convey certain image about oneself. These motivations also largely rely on the self-perceived stance towards separate opinions in general. The proponents of separate opinions view dissenting positively based on the separate opinions  being able to enrich the legal debate, being a sign of judicial independence, increasing the legitimacy of any given decision for it makes the decision more accurate of the real discussion behind it. The opponents of separate opinions mainly argue that showing the inability to speak in one voice undermines a court's legitimacy or the reputation of the dissenting judge. Moreover, judges seeking the appreciation from the general public or legal community may act in their personal interests instead of in the court's interests. Lastly, separate opinions come at collegiality costs and may harm the mutual relationships of judges.

In her theory of dissenting behavior, Wittig makes a sharp cut from the accounts coming mainly from the US, more specifically from the research on SCOTUS, and comes up with a model of separate opinions better suited for the civil law context of the CCC, the identification-disagreement model. Wittig argues that the traditional all have limited explanatory power as such and also do not fit within the civil law context, as judges therein are deciding in a different context, bound by different procedural rules, and, thus, given differing, sometimes broader, sometimes more limited, avenues to give way to their policy preferences or strategic considerations. We now discuss the identification-disagreement model in more detail.

## The Identification-Disagreement Model of Dissenting Behavior {#identification-disagreement}
Wittig introduces a non-formal model of separate opinions, the identification-disagreement model. Wittig amalgamates all the previously introduced potential motivations of judges for writing separate opinions into one cohesive and comprehensive model. The model is made up of two dimensions. The first dimensions of the model covers the disagreement level. The second dimension concerns the judges' stance and degree of self-identification of their role as a judge, Wittig terms this as a *norm of consensus*. Separate opinions are then "a function of a judge's identification with the norm of consensus and the level of disagreement of judges [@wittigOccurrenceSeparateOpinions2016, p. 74-75]. We base our theory on the identification-disagreement model and we use it to generate hypotheses for the CCC. Therefore, we now delve into the two dimensions deeper. We start with the norm of consensus and then we move on to the disagreement

### The norm of consensus
@calderiaTimeConsensualNorms1998, p. 876-877 define a norm as "a long-run equilibrium outcome, which underpins the interaction between individuals and reflects common understandings as to what is acceptable behavior in given circumstances." The norm of consensus in turn defines the level of dissent that is acceptable at any given court [@narayanConsensualNormHigh2005; @wittigOccurrenceSeparateOpinions2016, p. 75.]. Wittig's argument is two-fold. First, in civil law traditions unlike its US counterpart, the prevailing notion of the norm of consensus is that a court should not display disagreement. Second, the extent of adherence to the norm varies among judges,^[We conducted interviews with the justices of the third term of the CCC. Practically all of them more or less directly confirmed that they share the view that judges should not display dissent at a civil law court to a very varying degree.] depending on how they weight the costs and benefits they receive from following it [@wittigOccurrenceSeparateOpinions2016, p. 75.].

A dissonance between a proposed outcome for a case and any given judge's preferences are eventually bound to happen. In such a case, the judge can either express their sincere preferences by writing a separate opinion or they can adapt their behavior according to the norm of consensus and suppress the expression of her preferences. The second route has also been termed *dissent aversion* and theoretically fleshed out by @epsteinWhyWhenJudges2011. The decision of judge faced with such a conflict whether to attach a separate opinion or whether to avert their dissent is then a function of multiple potential utilities.

Wittig draws up three types of utility that dictate various levels of the adherence to the norm of consensus. Firstly, the intrinsic utility is maximized whenever a judge behaves in accordance with their true values and opinions, setting aside their strategic or political considerations. Secondly, expressive utility is harnessed when one displays individuality and counters the notion of conformism. Thirdly, the reputational utility arises when one adjusts their publicly displayed preferences to the expectations of others. Wittig argues that maximizing the former two forms of utilities in a situation of disagreement leads to separate opinions, whereas maximizing the reputational utility in such a situation gives way to the norm of consensus, as the judge would otherwise jeopardize the court's legitimacy as well as their reputation for not adhering to commonly accepted norms [@wittigOccurrenceSeparateOpinions2016, p. 76]. 

To some extent, we argue that even the third utility, the reputational utility, may lead to separate opinions insofar the individual reputation of a judge can in any way be linked to their non-conformity with the majority. An example that springs into mind is the late Justice Scalia, whose individual reputation among conservative circles would've been likely more jeopardized by siding with the liberal majority rather than with not adhering to the norm of consensus [@scaliaDissents1998]. The decision to dissent or to avert a dissent then is a result of an weighting between costs and benefits of these three types of utilities a judge derives from adhering or not adhering to the norm of consensus.

### Disagreement on the bench
A disagreement on a bench arises when the opinions on the matter diverge during a discussion and a judge has a reason to object the majority view. For example, case characteristics play an important role. Cases with more value-laden or controversial topics may give raise to more disagreement, similarly highly complex cases leave more space for disagreement. The sources of disagreement are seemingly manifold. To elucidate them theoretically, we rely on the case-space model.

The case-space model is a theoretical model developed in an attempt to model the idiosyncrasies of court decision-making, i.e. that a court is a body resolving disputes, cases [@landaDisagreementsCollegialCourts2007; @laxNewJudicialPolitics2011]. It differs from political science policy-space models in that that it incorporates the role of law. In the case-space model, the way to represent a case is by locating it in a n-dimensional space of of possible cases, the case-space. A case then denotes a legally relevant event that has occurred out of many that could have occurred. Put more simply, a case is a bundle of legally relevant facts. A legal rule then is a cut, a cut point when the case-space is one-dimensional, that divides the case-space. An individual disposition of a case is then the judgment of an individual judge of the case depending on their preference over legal rules (the cut point) and the location of the case in the case-space. In other words, if the case at hand falls to one side of the judge's cut point, then they vote for certain outcome, if it falls on the other side, they vote against it.

@landaDisagreementsCollegialCourts2007 draw from the case-space model multiple theoretical sources of disagreement. The first and clear source of disagreement among judges is that about facts. @landaDisagreementsCollegialCourts2007 conceptualize a couple of other sources of disagreement: which dimensions should be relevant under any given legal rule, disagreement about "thresholds" within dimensions and a couple of more sources of disagreement, which all can be summarised as a disagreement about the legal rule.

To this end, we simplify the model into two similar characteristics that concern the facts and the legal rule: case complexity and case controversy. Case complexity in our understanding refers to the number of legal issues that a case has touched upon. In line @landaDisagreementsCollegialCourts2007 we suppose that the more legal rules there is in play in any given case, the more room for disagreement about rules. We are also able to distinguish between different types of legal issues. Case controversy refers to the facts. There are subjects matters that we believe are more prone to disagreement. We discuss details in the [section](#operationalization) on operationalization.

### Strategic considerations
While we base our hypotheses mainly on Wittig's identification-disagreement model, we believe that strategic considerations as discussed by @epsteinWhyWhenJudges2011 can also come into play, albeit in a different institutional and regulatory setting than on SCOTUS. For example, workload varies across judge rapporteurs in the CCC context rather than across federal courts as in the US context. Therefore, we also base our one of our hypotheses on the strategic account of judicial decision-making. Rather than testing all potential sources of strategic considerations, we zero in on two specific sources of thereof, namely the role of collegiality costs and workload.

@epsteinWhyWhenJudges2011 address the issue of collegiality costs arising for a dissenting judge: "The effort involved in these revisions, and the resentment at criticism by the dissenting judge, may impose a collegiality cost on the dissenting judge by making it more difficult for him to persuade judges to join his majority opinions in future cases." Based on this theory, they predict and indeed empirically confirm that "dissents will be less frequent in circuits that have fewer judges because any two of its judges will sit together more frequently and thus have a greater incentive to invest in collegiality." Put simply, the researchers compare the dissent rates between courts with differing number of judges. 

The importance of leisure and workload on judicial decision-making has already caught attention of the empirically oriented legal scholarship. @clarkEstimatingEffectLeisure2018 have in a quasi-experimental difference-in-differences research design found that judges have preferences regarding their leisure, which then impacts their performance. They found that whenever a team of a US Federal Court judge's alma mater plays a basketball game, the time to process a decision increases for that respective judge and the quality of their work decreases. @brekkeThatOrderHow2023 found that the CJEU justices take into account their workload when issuing orders. In their paper on dissenting behavior of the SCOTUS justices, @epsteinWhyWhenJudges2011 found among others that judges strategically decide to write fewer dissenting opinions whenever the workload on their court increases. The evidence although overwhelming hasn't been one sided though. @songerExplainingDissentSupreme2011 found no relation between workload and dissent rate at the Supreme Court of Canada. 
The CCC's institutional set up also offers room for judges to adjust their behavior according to the workload they face. We now introduce the CCC, its institutional and procedural background, its powers as well as its composition.

# A brief primer on the CCC {#primer}
The CCC consists of fifteen justices, out of which one is the president of the CCC, two are vice presidents and twelve associate justices [following the terminology of @kosarConstitutionalCourtCzechia2020]. These fifteen justices are appointed by the president of the Czech republic upon approval of the Senate. The justices enjoy 10 years terms with the possibility of re-election; there is no limit on the times a justice can be re-elected. The three CCC functionaries are unilaterally appointed by the Czech president.

Regarding the competences, the CCC is a typical Kelsenian court inspired mainly by the German Federal Constitutional Court. The CCC enjoys the power of abstract constitutional review, including constitutional amendments. The abstract review procedure is initiated by political actors (for example MPs) and usually concerns political issues. Moreover, an ordinary court can initiate a concrete review procedure, if that court reaches the conclusion that a legal norm upon which its decision depends is not compatible with the constitution. Individuals can also lodge constitutional complaints before the CCC. Lastly, the CCC can also resolve separation-of-powers disputes, it can *ex ante* review international treaties, decide on impeachment of the president of the republic, and it has additional ancillary powers [for a complete overview, see @kosarConstitutionalCourtCzechia2020].

The CCC is an example of a collegial court. From the perspective of the inner, the CCC can decide in four bodies: (1) individual justices, (2) 3-member chambers (*senáty*), (3) the plenum (*plénum*), and (4) special disciplinary chambers. However, the 3-member chambers and the plenum play a crucial role. The plenum is composed of all justices, whereas the four 3-member chambers are composed of the associate justices. Neither the president of the CCC or her vice-presidents are permanents members of the 3-member chambers. Until 2016, the composition of the chambers was static. However, in 2016, a system of regular 2-yearly rotations was introduced, wherein the president of the chamber rotates to a different every 2 years.

In the chamber proceedings, decisions on admissibility must be unanimous, decisions on merits need not be, therefore, two votes are necessary.^[Which  enables the attachment of separate opinions] In the plenum, the general voting quorum is a simple majority and the plenum is quorate when there are ten justices present. The abstract review is one of the exceptions that sets the quorum higher, more specifically to 9 votes.

A judge rapporteur plays a crucial role [@chmelZpravodajoveSenatyVliv2017; @horenovskyProcessMakingConstitutional2015 study the large influence of the judge rapporteurs at the CCC]. Each case of the CCC gets assigned to a judge rapporteur. The assignment is regulated by a case allocation plan.^[The original term is *rozvrh práce*, which is usually translated as a *work schedule*, however, I borrow the term *case allocation plan* from @hamannGermanFederalCourts2019, p. 673] She is tasked with drafting the opinion, about which the body then votes. The president of the CCC (in plenary cases) or the president of the chamber (in chamber cases) may re-assign a case to a different judge rapporteur if the draft opinion by the original judge rapporteur did not receive a majority of votes. Unfortunately, the CCC does not keep public track of these reassignments.

The CCC allows for separate opinions. They can take two forms: dissenting or concurring opinions. Each justice has the right to author a separate opinion, which then gets published with the CCC decision. It follows that not every anti-majority vote implies a separate opinion, it is up to the justices to decide whether they want to attach a separate opinion with their vote. Vice-versa, not every separate opinion implies an anti-majority vote, as the justices can attach a concurring opinion. In contrast to dissenting opinion, when a judge attaches a concurring opinion, they voted with the majority but disagree with its argumentation.^[Which makes it difficult to, for example, conduct the same point-estimation with data on dissenting behavior of justices as @hanrettyDissentIberiaIdeal2012 has done on the Portugese and Spanish Constitutional Courts.] Unfortunately, it is difficult to conduct research on the dissent aversion because the voting is kept secret.

The room for the dissenting judge and the majority to address each other differs between the two bodies. Based on our qualitative research, there is less back and forth interplay between the judges, more akin to the SCOTUS context, and most of the communication is handled remotely in the panel proceedings, whereas the plenum meets regularly to discuss the cases in person. Despite that, procedurally speaking, the process of generating separate opinions is the same. In both cases, the rapporteurs are informed about the outcome of the vote, which is filed in the voting record. The separate opinion is then sent to the judge rapporteur before the decision is announced, as it cannot be added until after the announcement. It is important to note that judges have the possibility, not the obligation, to dissent. In other words, there is room for judges to give way to strategic considerations.

It may be concluded that the CCC takes after the American model of selection of justices, with the president of the republic and the upper chamber being in the spotlight, but is also a typical example of a Kelsenian specialised court with concentrated constitutional review. The CCC stands out in how powerful its constitutional review is, having attracted the power to review even constitutional amendments, thus, the CCC is a powerful player in the Czech political system. Despite its strength, it's not mandated to behave as one voice. Separate opinions are enabled by law in two forms, dissenting and concurring opinion. The leeway given to the CCC justices as to when they can attach a separate opinion opens up space for strategical considerations as well as consideration of the norm of consensus. Therefore, we believe the CCC to be an interesting subject of empirical study into the dissenting behavior of judges.

# Hypotheses
We draw multiple hypotheses from our theory. Firstly, following the identification-disagreement model, the probability of a separate opinion depends on judges' adherence to the norm of consensus and the level of disagreement. Therefore, the first two hypotheses are as follows:

**H~1~:** *The probability of observing a separate opinion is higher for judges with low norm-identification than for judges with high norm-identification.*

**H~2~:** *The probability of observing a separate opinion is higher for cases with a higher level of disagreement potential than in cases with a lower level of disagreement potential.*

The key dimension across which the extent of norm-identification may be theoretically expected to vary across judges is the professional background of the judges. The judges socialized within the judiciary are more likely to adhere to its values than those coming outside of it. In other words, CCC judges being appointed from the ranks of the judiciary are expected to adhere to the norm of consensus more than the rest. That applies especially strongly within the continental legal system, in which voicing a disagreement is usually prohibited within the ordinary judiciary. In Czechia, apart from the CCC and the Supreme Administrative Court, no separate opinions are even legally allowed. Therefore, the CCC justices coming out of the judiciary are socialized in an environment, in which voicing a dissent is not even allowed by law. On the other hand, lawyers or justices recruited from the ranks of the politicians are not socialized from within the judiciary and are used to defending their own opinion, or at least they are not required to hide their disagreement. However, we also theoretically expect that the difference between the professions should wane over time as the non-judicial justices get accustomed to the role of a CCC judge. Our hypothesis 3 suggests: 

**H~3~:** *The closer the date of the decision to the date at which the judge entered, the larger the difference between the probability of observing a separate opinion of justices coming from judiciary and of justices coming outside judiciary.*

According to @epsteinWhyWhenJudges2011: "[t]he economic theory of judicial behavior predicts that a decline in the judicial workload would lower the opportunity cost of dissenting [...]." Using the language of the identification-disagreement model, we believe leisure to be an example of the individual utility a judge may consider. Contrary to Wittig, we believe individual utility may also pull the other way: against a separate opinion. Therefore, our hypothesis 4 suggests:

**H~4~:** *The higher the workload of a judge, the lower the probability of them attaching a separate opinion.*

Moving on to the collegiality costs, we test whether judges that are at the start of their term, and thus are aware that they will "sit together more frequently" invest in collegiality by averting separate opinions and whether when their term draws to an end, they give way to their disagreement. This presumes that the outlook of sharing the 10 year term with your colleagues at the beginning of judges' terms increases the collegiality costs of dissenting, whereas at the end of their terms, the collegiality costs decrease with the end of the shared term looming on the horizon.

**H~5~:** *The closer the date of the decision to the date at which the judge entered the office, the lower probability of a separate opinion, whereas the closer the date of the decision to the date at which the judge left the office, the higher the probability of a separate opinion.*

# Empirical analysis {#empirics}
We now move onto the empirical part of our paper. First, we describe the data we collected. Second, we explain how we operationalized the variables included in our hypotheses. Third, we explain our identification strategy. Lastly, we present the results of our model.

## Data description
The data is based on the CCC dataset [@paulikCzechConstitutionalCourt2024], which contains all decisions published by the CCC since its foundation, complete text corpus as well as plenty of metadata. The whole dataset is available at [Zenodo](https://zenodo.org/records/10591110). We narrow our cases to all plenum decisions and to all 3-member chamber decisions on merits up until the end of 2022. The admissibility decisions of the 3-member chambers must be made unanimously, concurring decisions therein are a rarity.^[On top of that out of the 39 separate opinions in admissibility 3-chamber decisions, 25 of that are a copypasta from justice Jan Filip and 6 are a copypasta from justice Josef Fiala in alike cases. Thus, there is only a few left for a meaningful analysis. The class imbalance of the remaining ~10 decisions would be too large against all chamber decisions on admissibility] Fig. \@ref(fig:separate-opinion-ratio) confirms our intuition about the lack of variance among the admissibility decisions.

```{r separate-opinion-ratio, fig.cap = "Percentage of Decisions with at least one Separate Opinion"}
separate_opinion_ratio
```

We skip the first decade of the CCC as the data on it are rather incomplete and inconsistent: For example, very few decisions contain the information on the composition in the text and many do not contain the name of the dissenting judge at all. We also limit our analysis until the end of 2022 as the CCC entered its 4th term in 2023 and started to undergo a complete personal change. 

The final dataset for analysis contains `r toString(length(unique(data$doc_id)))` decisions of the CCC. Table \@ref(tab:overalltable). reveals that out of them, `r overall_table[overall_table$formation == "Chamber", "ratio_total"]` are decisions by the 3-member chambers on merits, around `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "merits", "ratio_total"]` are plenum decisions on merits and the remaining `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "admissibility", "ratio_total"]` are plenum decisions on admissibility. At least one separate opinion contain `r overall_table[overall_table$formation == "Chamber", "ratio_dissent"]` decisions out of the 3-member decisions, `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "admissibility", "ratio_dissent"]` decisions of the plenum decisionson admissibility, and `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "merits", "ratio_dissent"]` decisions of the plenum decisions on merits.

```{r overalltable}
overall_table %>%
  kable(
    caption = "Summary statistics for the whole dataset", 
    col.names = c("Formation", "Grounds", "Count", "%", "% with at least One Separate Opinion")
  ) %>%
  row_spec(0,bold=TRUE) %>%
  kable_styling()
```


## Operationalization {operationalization}
We now discuss how we operationalized the variables included in our hypotheses. We start with the dependent variables, then we move on to the explanatory variables and finish with control variables.

### Dependent variable: a separate opinion
We conceptualize our dependent variable as the information whether a justice attached a separate opinion to a decision or not.^[Unlike Wittig we do not call our dependent variable a judge's vote, as that refers to a slightly different thing within the CCC context. A judge may vote against the majority opinion but since they are not mandated to write a separate opinion, these do not necessarily overlap. Similarly, a judge may vote for a disposition of a case and still attach a concurrence separate opinion.] Our dependent variable *separate opinion* is thus a dummy variable that has two categories: either a justice did attach a separate opinion (1) or she did not (0) in any given case .

We do not distinguish between a concurrence and a full dissent. The reason is two-fold: practical and theoretical. Theoretically, the difference between the two lays only in the disposition of the case. The justices may equally disagree on the interpretation of legal rules, thus, in the case-space model terms [@landaDisagreementsCollegialCourts2007; @laxNewJudicialPolitics2011], the judge cut points in any given case differ even when a justice attaches "only" concurrence. The difference is that in the cases containing concurrence, the case facts may have completely accidentally fallen on the same side both of the concurring judge as well as the majority, whereas in the cases containing a dissenting opinion, the case fell in between the cut points. We do not consider this phenomenon as theoretically interesting. Practically, it was impossible to distinguish between the two categories as some decisions simply do not contain the information whether a separate opinion was a concurring or dissenting opinion.

### Explanatory variables
#### Disagreement potential {-}
From the theoretical perspective it may be expected that the potential for disagreement varies across cases. In some cases, the disagreement potential is higher and, thus, the probability of a separate opinion is higher than in the cases with lower potential for disagreement. More specifically, based on our theory we expect the potential for disagreement to be captured by two characteristics of any given case: (1) its legal complexity and (2) its controversy or in terms of @epsteinMeasuringIssueSalience2000 issue salience.

##### Complexity {-}
As @corleyPuzzleUnanimityConsensus2013 argue and empirically measure, complexity of a case leads to less certainty and more ambiguity for the justices, which leads to a higher probability of disagreement. The authors define legal complexity as the number of legal issues a case has to address. True to the @corleyPuzzleUnanimityConsensus2013 study, our operationalization of complexity relies on the assumption that the more legal issues there are in any given case, the higher the number of references to other laws and caselaw in the text of the corresponding decision. What's more, we are able to distinguish between legal issues at the constitutional and ordinary level.

The variable *concerned acts* captures the number of concerned ordinary legal acts on the legal-act level, the variable *concerned constitutional acts* captures the number of articles of the constitutional legal acts (mostly the Czech Constitution and the Charter of Fundamental Rights and Freedoms) and variable *caselaw* captures the number of citations to its own caselaw. The information on the first two variables is based on the metadata provided by the CCC, the last is based on the regular expressions search of the text of the decisions.

```{r}
corr_acts %>%
  as_tibble() %>%
  mutate(
    term = c("1. # of Ordinary Acts", "2. # of Constitutional Articles", "3. # of CCC References")
    ) %>%
  kable(
    caption = "Correlations between Independent Variables of Case Complexity",
    col.names = c("Variable", "1", "2", "3")
  ) %>%
  row_spec(0,bold=TRUE) %>%
  kable_styling() %>%
  footnote(
    general = "Correlation was calculated using the Spearman Correlation Coefficient.",
    general_title = "Note.",
    footnote_as_chunk = TRUE
    )
```

The first two variables we believe to be sufficiently different from each other: a typical right to fair proceedings case may concern only one constitutional article (the article 36 of the Charter) but many legal acts, whereas a typical separation of powers case concerning the Parliament concerns many constitutional articles but only few ordinary laws. On the other hand, the number of (concerned) constitutional acts and references to the CCC caselaw may be correlated. We therefore run diagnostics, which reveal that the citations to CCC caselaw and the number of references to constitutional acts are rather correlated. Because we believe the number of cited constitutional articles to better reflect the number of legal issues than the number of references to CCC case-law, in our final model we exclude the number of references variable.

##### Controversy/Issue salience {-}
In a similar vein, certain typically value-laden issues may generate more disagreement even if they raise only one or few legal questions. @epsteinMeasuringIssueSalience2000 distinguish between two types of issue salience. Retrospective issue salience is how a particular issue is viewed in hindsight, regardless of whether the actors at the time regarded it so. Contemporaneous salience than captures the stance of the relevant actors, the CCC justices in our cases, on the issue. While we are interested in the latter, it is not obvious as to how exactly can such an information be captured. @epsteinMeasuringIssueSalience2000 reject all the typically employed proxies for issue salience such as the number of articles in law review that a case generated or the number of citations a case generated. Instead, they propose a novel unbiased transparent measure of issue salience: whether a SCOTUS decision made it to the front-page of the New York Times newspaper. Unfortunately, practically all of the measures are hard to transfer to a non-US context. We cannot exactly pinpoint a similar newspaper as New York Times in Czechia. Therefore, we need to rely on another measure of issue salience/case controversy.

In her thesis, @wittigOccurrenceSeparateOpinions2016 selected a number of subject matters she deemed as particularly controversial. In the CCC context, there are a number of issue that have been deemed as controversial by the Czech legal scholarship. Typically, the restitution cases or cases concerning fundamental human rights have been coined as rather controversial. The CCC dataset contains a table *subject_matter*, which contains the subject matter of any given case (a discrimination case, a separation of powers case and alike) and information on the area of constitutional law that the case pertains (the right to fair trial, the right of freedom of speech).

Following Wittig's method, we coded the topics as controversial when they concern any of the following subject matters: discrimination, expropriation, restitution, the property of church, sexual orientation, the protection of consumer, fundamental human rights, social and cultural rights, the right of property, the freedom of speech, and the separation between the church and state. Therefore, the variable *controversial* is a dummy variable, which takes the value of 1 when the given case falls in any of the previously mentioned categories or it takes the value 0 when it does not.

#### Norm-identification {-}
Secondly, our theory generated a second expectation: a relationship between adherence to the norm of consensus and the likelihood of a separate opinion. The adherence to the norm of consensus depends on self-identification and on self-perception of the judges: "being part of the system of the judiciary influences identification with the organization's goals and it shapes considerably the judges' perception of how to behave." [@wittigOccurrenceSeparateOpinions2016, p. 100]. Theoretically, we expect the adherence to the norm of consensus to depend on one's career choices. The actors socialized within the judiciary and its values are more likely to adhere to them than their peers who entered the CCC from the legal practice, from politics, or from anywhere outside the judiciary [@wittigOccurrenceSeparateOpinions2016, p. 84-87.].

```{r age-distribution, fig.cap = "Kernel density of justices' ages at the end of their terms."}
age_distribution
```

Wittig confirms the theoretical intuition empirically as well but she operationalizes the norm-identification as the justices' career choice after their term. The justices that chose to stay within judiciary or go back to being scholars are expected to more strongly identify with the norm of consensus, whereas the justices' that made different career choice are less likely to identify with the norm of consensus. Unfortunately, such a measure does not fit the CCC context. As the table \@ref(fig:age-distribution), CCC justices start their term usually at the end of their career and a considerable part of them ends their term in their 70s, well past the retirement age in Czechia. Given that the retirement age of judges 65 years of age, we expect majority of the CCC justices to simply retire after their terms.

```{r profession-distribution, fig.cap = "The distribution of professions of the CCC justices across its 3 terms."}
professions
```

Therefore, instead of operationalizing the norm-identification as the profession after their term, we operationalize it as the last profession before they entered the CCC. Viewed this way, we presume that the last profession that a justice had held before entering an office had the highest impact on their self-identification as a judge. Overtime we expect the effect to wane off. 
@garoupaDisagreeingPrivateDissenting2022 have found evidence that the individual decision to attach a separate opinion of the judges on the Spanish Council of State depends on the justice's professional background. The division does not cut across as we would theoretically expect: only former politicians seem to be less likely to attach a separate opinion, while former judges and lawyers seem to be more likely to attach a separate opinion.

Quite handily, the CCC dataset contains such an information on the last profession before the CCC justices took up their office. In the dataset, the variable *judge_profession* contains the information on the justices' previous career choices and can take up the values of judge, scholar, politician, or lawyer. The table  \@ref(fig:profession-distribution) reveals that the distribution of the professions has changed across time. For our purposes, we flatten the original variable into 2 levels. The variable *profession* is coded either as "within" if the justice previously served as a judge or "outside" if they came from the other three professions.

#### Time in office {-}
The effects of the CCC justices' profession should according to our theoretical expectations vary depending on the time the judges have spent in their office. The closer to them taking up the office, the effects of their preceding profession should be more pronounced. Similarly, according to the collegiality costs hypothesis, the dissenting behavior of justices should vary across time. Namely, justices are expected to dissent less likely at the beginning of their terms, as the collegiality costs are higher, and to dissent more likely at the end of their terms, as the collegiality costs of the disagreement are lower. We capture the *time in office* variables as the number of months a justice has left until their end of mandate at the date of the decision. That allows us to account both for the collegiality costs hypothesis as well as the difference between professions hypothesis.

#### Workload {-}
Similarly to the study of @brekkeThatOrderHow2023 on workload at the CJEU, we operationalize *workload* as the number of pending unfinished cases that any given judge has in the moment of any given decision as a judge rapporteur. We firstly mined the compositions of panels as well as the plenary from the text of the decision. We then calculated the number of unfinished cases each judge had at the time of any given decision as a judge rapporteur using the date of submission and of decision of a case. The mean number of pending cases of any justice as a judge rapporteur is `r descr_workload$numeric.mean[1]`, with quite a large standard deviation of `r descr_workload$numeric.sd[1]`. The highest number of pending cases in the dataset is `r descr_workload$numeric.p100[1]` in the case of the justice Radovan Suchánek, whose work ethic has already become a topic of scholarly discussion [@chmelCoOvlivnujeUstavni2021].

```{r workload}
descr_workload %>% 
  kable(
    caption = "Summary statistics for the workload variable", 
    col.names = c("", "Mean", "SD", "P0", "P25", "P50", "P75", "P100"),
    digits = 0
  )
```

One looming issue arises that needs to be addressed and may introduce bias. Not every unfinished pending case is the same. Different judges may cope with the increasing workload differently. Czech scholars have argued that deciding cases on admissibility, which are typically shorter and require a rather concise argumentation, may be such a coping mechanism of overburdened judges: it allows them to expedite decisions faster and with less effort. Thus, the same number of unfinished pending cases of a judge that chooses the strategy of offloading their workload burden into decisions on admissibility may not carry the same workload as with a judge that opts for decisions on merits more often. As a solution, we multiply the workload by a ratio of unfinished cases on merits to admissibility.

<!-- ### Control Variables -->
<!-- Lastly, we control for potential confounding variables. A confounding variable is such that (1) has an effect on treatment status and (2) has an effect on the outcome over and above its effect on the treatment status. Not controlling for confounding variables causes an omitted variable bias [@angristMostlyHarmlessEconometrics2009; @angristMasteringMetricsPath2014; @buenodemesquitaThinkingClearlyData2021]. Therefore, we need to carefully consider which additional features may induce bias rather than throw in as many additional variables as possible. -->

<!-- There are certain institutional and procedural features that have been traditionally controlled for in the studies, in which the observation is a court decision or a judge's vote. -->

<!-- #### Grounds {-} -->
<!-- At the FCC, procedural specifics play a central role in a judge’s opportunities to deliver a separate opinion. While several courts such as the US Supreme Court have complete control over their docket, other courts are compelled to hear a case if it meets certain procedural conditions. The FCC employs such a system. However, since research on courts with admissibility requirements is scarce and not much data are available, this aspect has not found its way into empirical studies, yet. Being able to draw on the CCDB, the analysis to follow accounts for this procedural rule for the first time. At the court, the examination of a case consists of two parts. First, the judges scrutinize the admissibility of a case. Only if a plaintiff’s request meets certain formal and substantial criteria the court proceeds to examining the questions on the merits. Thus, the content of a case may be extremely complex, highly controversial, and the judges split over the topic; but if a plaintiff’s application is inadmissible, neither the high degree of controversy nor the discordance within the senate will come into effect. Nevertheless, it should be noted that although implying clear criteria, the check for admissibility still provides some room for discretion. Therefore, rejecting a request on grounds of inadmissibility does not exclude separate opinions per se, yet it curtails it. Still, in order to cover the conditions under which a decision is made, the procedural confinement of admissibility needs to be included in the analysis. In the dataset the variable Admissibility comprehends the information whether a case was admissible or not. However, these data are collected in the CCDB on the proceedings level. Therefore, I collapsed it in order to include it in an analysis on the case level. Whenever one of the proceedings was at least admissible in parts, the variable was coded 1 as even a partial admissibility leads to examination on the merits. 2 stands for inadmissibility and 3 for those three categories that cover decisions in which for different reasons the court could not examine the admissibility. -->

## Identification Strategy
The hypotheses are tested by fitting a generalized linear model estimating the probability of a judge attaching a separate opinion with the dependent variable following a binomial distribution. 

``` {r model-selection}
model_selection %>%
  kable(caption = "Descriptive statistics for the formation clusters",
        col.names = c("Units", "Mean Observations per Unit"))
  
```

When identifying our model, we stood before a research choice that has been addressed differently by different researchers. As other studies on judicial decision-making relying on observational data, the data on judicial-decision making offers multiple potential clustering variables to include fixed effects on. In their study of effect of french language on the performance of the judges at the CJEU, @cheruvuHowInstitutionalConstraints2019 used fixed effects in their model the effects both on the individual judges as well as on the panel level. The reasoning behind fixed effects clustered on the formation were that it "[s]cholars argue the number of judges sitting on a chamber is a proxy for a cases’ salience as more important cases tend to be assigned to larger chambers (e.g. @kelemenPoliticalFoundationsJudicial2013; @larssonJudicialIndependencePolitical2016). Including chamber fixed-effects in my models addresses both heterogeneity in the collegial decision-making process of different chambers of judges and implicitly controls for the number of judges hearing a case." @clarkEstimatingEffectLeisure2018 in their paper studied the effect of leisure on judicial performance.^[Also measured as time for a judge to decide a case similarly to the @cheruvuHowInstitutionalConstraints2019 study.]. They included the the author fixed effects in their model.^[Which would correspond to the judge rapporteur in our case.] @brekkeCJEUDatabasePlatform2023 in a study on the usage of orders at the CJEU built a logistic regression model with subject matter fixed effects.

Because in our study, we are interested in the between variance of individual judges as well as the subject matters, it does not make sense to include fixed effects of either cluster. In contrast to that, we are not interested in the so-called panel effects, especially the difference between the 15-member plenum  and the 3-member panels could produce bias. The more controversial cases or more legally complex cases are typically assigned to the plenum and the occurrence of separate opinions is higher in the plenary decisions. Moreover, the behavior of the 3-member panels varies as well. It thus makes the most sense to include panel fixed effects as that would allow us to control for a potentially potent source of bias. In our model, the variable *formation* captures the information on the 3-member panels, whose composition is relative unstable especially after the introduction of system of rotations of justices in 2016, therefore the information is coded as the exact composition of any given 3-member panel, and the information whether the case was decided by the plenum, whose composition is relatively stable within the two terms.

Moreover, our data includes two terms of the CCC. Its second decade roughly between 2003 and 2013 and its second decade between 2014 and 2023. The composition of the CCC has completely changed between the two terms and so has, we expect, changed the behavior of the court as such. Rather than including year fixed effects, we opt for the term fixed effects. If the majority of judges sitting in a panel came from the first term, the variable *term* is then coded as "1st", and if the majority of justices sitting in a panel came from the second term, the variable takes up the value "2st".

## Results and discussion

<!-- **H~1~:** *The probability of observing a separate opinion is higher for judges with low norm-identification than for judges with high norm-identification.* -->

<!-- **H~2~:** *The probability of observing a separate opinion is higher for cases with a higher level of disagreement potential than in cases with a lower level of disagreement potential.* -->

<!-- **H~3~:** *The closer the date of the decision to the date at which the judge entered, the larger the difference between the probability of observing a separate opinion of justices coming from judiciary and of justices coming outside judiciary.* -->

<!-- **H~4~:** *The higher the workload of a judge, the lower the probability of them attaching a separate opinion.* -->

<!-- **H~5~:** *The closer the date of the decision to the date at which the judge entered the office, the lower probability of a separate opinion, whereas the closer the date of the decision to the date at which the judge left the office, the higher the probability of a separate opinion.* -->

The results reveal quite an interesting trend at the CCC. We tested the model under more specifications. The full model presented the best fit. However, while excluding the disagreement potential variables produced a significant drop in AIC, excluding the norm-identification did not. Therefore, the norm-identification variables do not seem to carry too much explanatory power regarding the dissenting behavior of CCC justices. 

```{r, warning = FALSE, echo = FALSE}
modelsummary::modelsummary(list("Full Model" = model_me_formation,
                                "Disagreement" = model_me_disagreement,
                                "Norm" = model_me_identification),
                           output = "latex",
                           estimate = "{estimate}{stars}",
                           statistic = "({std.error})",
                           stars = TRUE,
                           title = "Results from the Logit Model")
```

Moving on to the interpretation of the results, our results do not support the Hypothesis 1. Unlike at the GFCC, we do not find evidence for the norm-consensus operating according to our theoretical expectations. Firstly, adding the interaction term between the professional background and the time in office does not improve the model fit. It does not carry much explanatory power as an ANOVA test fails to reject the null hypothesis of no difference. Secondly, neither the separate variables of norm-identification nor the interaction term are near statistical significance. That does not however lead to us rejecting the identification-disagreement as a theoretical model. We believe it to be a clear theoretical model that can guide further empirical inquiries. It may just be the case that the proxy of the justices' profession may not be wholly accurate or that even if it was there may simply be currently no norm of consensus operating at the CCC.

In contrast to that, we find a clear support for Hypothesis 2: all the disagreement potential variables are positively and significantly correlated with the probability of a judge attaching a separate opinion. No matter the model specification, we always found a clear statistical, substantive and positive relationship between the probability of a judge attaching a separate opinion and the number of constitutional or ordinary acts concerned.

The same applies to the controversial value-laden topics. Because we are aware of the potential issue with "arbitrarily" selecting a couple of potential subject matters, we decided to test the robustness of our *controversial* variable. To do so, we conducted a placebo test. We sampled 10 placebo samples of different subject matters (n = `r length(controversial_topics)`) and fitted the same final full mixed-effects model with the *controversial* variable being replaced by the placebo. After rerunning the placebo test multiple times with various samples of different seeds, we found that only sometimes would at most one placebo sample out of the 10 be statistically significant, the remaining 9 or 10 samples were not.  While such a conclusion may not come as a complete surprise, as it is perfectly aligned with theoretical expectations, we provide a firm evidence for it.

``` {r placebo-test}
placebo_models %>%
  select(-statistic) %>%
  kable(caption = "The result of 10 models fitted with different placebo sample from the subject matter variable.",
        col.names = c("", "Estimate", "SE", "p.value"),
        digits = 2)
```


Our results also support for Hypothesis 4. Higher workload of a judge decreases their probability of attaching a separate opinion. It thus appears that neither the claims of @epsteinWhyWhenJudges2011 are not unwarranted and may carry over to other contexts. The results thus reveal that judges of the CCC give way to strategical considerations regarding their leisure. If they're overburdened with work, they reduce the load elsewhere, namely in the additional burden of writing separate opinions.

Lastly, we do not find evidence for the so-called freshman's effect. The importance CCC justices attach to collegiality costs of a separate opinion seem to remain consistent across time as the time in office of a justice does not carry explanatory power in our model.

# Conclusions {#conclusions}

\vspace{30pt}

# References {-}

<div id="refs"></div>

# (APPENDIX) Appendix {-}

# Appendix