---
output: 
  bookdown::pdf_document2: 
    toc: false
    fig_caption: yes
    number_section: true
title: "Disagreement on a bench: an empirical analysis of dissent at the Czech Constitutional Court"
author:
- name: ANONYMIZED
- name: ANONYMIZED
anonymous: FALSE
removetitleabstract: FALSE
abstract: "This article examines the factors influencing dissenting opinions by judges at the Czech Constitutional Court (CCC). We build on the disagreement-identification model and the strategic accounts of dissenting behavior. Our findings do not support the existence of a strong norm of consensus operating at the CCC. However, the complexity of a case, measured by the number of CCC caselaw citations involved, is positively correlated with the probability of a dissenting opinion. Similarly, cases concerning controversial topics are more likely to generate dissents.  A placebo test strengthens this finding by demonstrating that randomly chosen, non-controversial topics have minimal impact on dissenting behavior. When facing a high workload, judges are less likely to write separate opinions. The effect of voting blocs in the plenary proceedings seems to carry over to the 3-member chamber proceedings. The results also reveal that judges make strategic considerations. When facing a high workload, judges are less likely to write separate opinions, suggesting they prioritize workload management. Lastly, CCC judges seem to take into account collegiality costs as their are more likely to dissent at the end of their terms than at their start."
thanks: "This article was created as part of the Specific University Research (SVV) project of GRANT ANONYMIZED"
keywords: "empirical legal research, courts, dissents, judicial behavior, political science, regression analysis"
date: "`r format(Sys.time(), '%B %d, %Y')`"
link-citations: yes
linkcolor: blue
geometry: margin=1in
fontsize: 11pt
doublespacing: TRUE
endnote: no
# pandocparas: TRUE
sansitup: FALSE
mainfont: Times New Roman
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{float}
bibliography: "`r rbbt::bbt_write_bib('bibliography.bib', overwrite = TRUE)`"
---
```{r setup, include=FALSE}
library(patchwork)
library(tidyverse); theme_set(theme_minimal())
library(ggplot2)
library(kableExtra)
library(modelsummary) 
library(bookdown)
library(fixest)
library(tidymodels)

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.asp=0.8, fig.pos = "H", out.extra = "", fig.align = 'center')

load("../data/descriptive_statistics.RData")
load("../data/model_results.RData")
```

# Introduction
"I don't like them [separate opinions]. (...) Because I am a routine judge and I am of the opinion that when a collegiate body makes a decision, a person X shouldn't further comment on it, just because they were of a differing view. It is undermining the authority of that court." The previous quote has been voiced by one of the Czech Constitutional Court ("CCC") judges. It expresses one of many possible stances a judge may take towards exercising their right to dissent that's been shaped and influenced by their view of the role of a CCC judge and the CCC itself, a concept we later coin a norm-identification. The possible aspects influencing the judges' decision whether to dissent or not is manifold and has encompassed political considerations [@hanrettyJudicialDisagreementNeed2015], strategic considerations [@epsteinWhyWhenJudges2011], or institutional-systemic consideration [@garoupaSpurredLegalTradition2020].

This paper presents an empirical study of dissenting opinions at the CCC. We explore the factors that influence a judge's decision to dissent stemming from their strategic considerations and stemming from the identification-disagreement model laid out in @wittigOccurrenceSeparateOpinions2016. We consider whether judges react to their workload, to their previous professional history, to their colleagues, or to the characteristics of the case itself. We do not consider the institutional-systemic influences such as political fragmentation.

In the Czech context, empirical legal scholarship on the dissenting behavior has been sparse. It has focused mainly on analyzing the presence of implicit voting blocs in the last term of the CCC. Because the votes are not known, in conducting a network analysis, the researches have relied on the information, which judge dissented in which decision. The research has revealed that the third period of CCC between 2013-2023 is rather polarized and that there are two big dissenting coalitions of judges that clash against each other [@chmelCoOvlivnujeUstavni2021; @smekalMimopravniVlivyNa2021; @vartazaryanSitOvaAnalyza2022]. The Czech studies include in their analysis only cases with a separate opinion and do not compare it against decisions without it.

In the civil law context, @kelemenJudicialDissentEuropean2017 offers a mainly theoretical comparative overview of the various regimes of dissenting behavior across European courts. Hanretty has made use of dissenting behavior of Spanish and Portuguese judges to conduct a point estimation of location of the "political" position of the judges [@hanrettyDissentIberiaIdeal2012] or to investigate what does a dissent reveal about the dimension across which the disagreement runs on Estonian Supreme Court [@hanrettyJudicialDisagreementNeed2015] and the British Law Lords [@hanrettyDecisionsIdealPoints2013]. @garoupaSpurredLegalTradition2020 studied the influence of political fragmentation on the dissenting behavior of Slovenian and Croatian judges. Most importantly, @wittigOccurrenceSeparateOpinions2016 has put forward and empirically tested on the German Federal Constitutional Court a theoretical *disagreement-identification* model of dissenting behavior tailored at the civil law legal systems.

We contribute to the existing literature. Unlike the Czech studies, we consider a wider range of decisions of the CCC. We also put into use and test the disagreement-identification model. We find that in line with the identification-disagreement model, the disagreement potential of any given case is positively correlated to the likelihood of a separate opinion. In contrast to that, we do not find evidence that the adherence to the norm of consensus varies across professional background of the judges. We also find that judges' workload is correlated with their likelihood of attaching a separate opinion. Lastly, we do not find evidence that the dissenting behavior varies depending on the time elapsed between the start of judges' terms and the date of a decision.

We believe the CCC to be an object worthy of study. First, the appointment procedure of the CCC judges may be compared to the SCOTUS, whereas its role within the constitutional system is akin to the European constitutional courts. That allows us to build on and to contribute to the common law as well as civil law oriented empirical legal research. Secondly, the CCC's decisions have an impact as it has been vested with large amount of competences: it may review laws in abstract as well as in concrete proceedings initiated by an individual complaint.  On top of that, the CCC has broadened its power and may now review even constitutional amendments and historically has not been afraid to step in into politically salient cases. In the *Melčák* case, the CCC entered a heated political battlefield by annulling a constitutional amendment, which shortened the term of the Chamber of Deputies, and sparked a drawn out constitutional crisis.^[The English version of the decision is available at: https://www.usoud.cz/en/decisions/2009-09-10-pl-us-27-09-constitutional-act-on-shortening-the-term-of-office-of-the-chamber-of-deputies.] Thirdly, separate opinions have been allowed for at the CCC since its inception, their practice has been firmly established. The main drawback of the CCC institutional setup is the inability to measure the role of political preferences of judges due to the lack of information on judges' votes [@martinDynamicIdealPoint2002; @hanrettyDissentIberiaIdeal2012] and the lack of variance in the nomination process.^[As explained bellow, practically all judges of one term have been nominated and appointed by the same President and Senate.]

Our article proceeds as follows. In [section 2](#theory), we discuss the strategical and the identification-disagreement accounts of dissenting behavior. Therein, we draw hypotheses based on the general theory applied to the specific institutional setup of the CCC. In [section 3](#primer), we introduce the CCC. We discuss its institutional setup, its procedures and, most importantly, the rules concerning separate opinions.  In [section 4](#method), we discuss our method, our model, and the operationalization of variables.  In [section 5](#empirics), we empirically test the hypotheses and we discuss the results. [Section 6](#conclusion) concludes. 

# Theories of Dissenting Behavior {#theory}
## Strategic Account of Dissenting Behavior
@epsteinStrategicRevolutionJudicial2000 argue that judges as "(1) social actors make choices in order to achieve certain goals, (2) social actors act strategically in the sense that their choices depend on their expectations about the choices of other actors, and (3) these choices are structured by the institutional setting in which they are made."  Instead of holding judges to pursue political policy oriented goals, the judges' self-interest in terms of career progression, higher income, more leisure, or lesser workload takes the spotlight [@epsteinStrategicRevolutionJudicial2000]. In their empirical study on dissenting behavior on the Supreme Court of the USA ("SCOTUS"), @epsteinWhyWhenJudges2011 posit that "a potential dissenter balances the costs and benefits of issuing a dissenting opinion." and that judges have "leisure preferences, or, equivalently, effort aversion, which they trade off against their desire to have a good reputation and to express their legal and policy beliefs and preferences (and by doing so perhaps influence law and policy) by their vote, and by the judicial opinion explaining their vote, in the cases they hear."

The utility of a dissenting opinion may take form of the potential to undermine the majority opinion when the dissent is influential and the enhanced reputation that the judge enjoys as the dissenting opinion may be cited in the future by other judges or publicly analysed by legal scholars. The costs take up two forms. @epsteinWhyWhenJudges2011 argue that the judges strategically take into account collegiality costs, which are lower at courts that sit in larger chambers, whereas they are bigger at courts that decide in smaller chambers as the judges have to spent time in a smaller circle of their colleagues. Moreover, they predict that the judges may reap benefits of averting a dissent whenever they face a high workload so that they free up their hand to take care of more pressing work. @epsteinWhyWhenJudges2011 find that the US judges indeed take into account the reputation utility, the collegiality costs as well as their workload.

@epsteinWhyWhenJudges2011 address the issue of collegiality costs arising for a dissenting judge: "The effort involved in these revisions, and the resentment at criticism by the dissenting judge, may impose a collegiality cost on the dissenting judge by making it more difficult for him to persuade judges to join his majority opinions in future cases." Based on this theory, they predict and indeed empirically confirm that "dissents will be less frequent in circuits that have fewer judges because any two of its judges will sit together more frequently and thus have a greater incentive to invest in collegiality." We presume that at the beginning of judges' term, they are aware that they will "sit together more frequently" and that the outlook of sharing the 10 year term with your colleagues at the beginning of judges' terms increases the collegiality costs of dissenting, whereas at the end of their terms, the collegiality costs decrease with the end of the shared term looming on the horizon. We test whether judges that are at the start of their term invest in collegiality by averting separate opinions and whether when their term draws to an end, they give way to their disagreement. We draw our first hypothesis:

**H~1~:** *The closer the date of the decision to the date at which the judge left the office, the higher the probability of a judge attaching a separate opinion in that case.*

According to @epsteinWhyWhenJudges2011: "[t]he economic theory of judicial behavior predicts that a decline in the judicial workload would lower the opportunity cost of dissenting [...]." @clarkEstimatingEffectLeisure2018 have in a quasi-experimental difference-in-differences research design found that judges have preferences regarding their leisure, which then impacts their performance. They found that whenever a team of a US Federal Court judge's alma mater plays a basketball game, the time to process a decision increases for that respective judge and the quality of their work decreases. @brekkeThatOrderHow2023 found that the CJEU judges take into account their workload when issuing orders. @engelMannaHeavenJudges2020 found in a quasi-experiment in Israel that reduction in judges' workload affects the way they decide. @epsteinWhyWhenJudges2011 found among others that judges strategically decide to write fewer dissenting opinions whenever the workload on their court increases. The evidence although overwhelming hasn't been one sided though. @songerExplainingDissentSupreme2011 found no relation between workload and dissent rate at the Supreme Court of Canada. Therefore, our hypothesis 2 suggests:

**H~2~:** *The higher the workload of a judge at the time of a decision, the lower the probability of them attaching a separate opinion.*

<!-- @wittigOccurrenceSeparateOpinions2016 in her study on separate opinions at the Federal Constitutional Court of Germany ("GFCC") summarizes the potential motivations for judges to attach a separate opinion and, thus, to acquire additional costs: (1) potential of impacting future caselaw, (2) moral obligation to distance oneself from a decision that contradicts their values, (3) to convey certain image about oneself. These motivations also largely rely on the self-perceived stance towards separate opinions in general. The proponents of separate opinions view dissenting positively based on the separate opinions being able to enrich the legal debate, being a sign of judicial independence, increasing the legitimacy of any given decision for it makes the decision more accurate of the real discussion behind it. The opponents of separate opinions mainly argue that showing the inability to speak in one voice undermines a court's legitimacy or the reputation of the dissenting judge. Such a view is perfectly in line with the quote from the introduction. Moreover, judges seeking the appreciation from the general public or legal community may act in their personal interests instead of in the court's interests. Lastly, separate opinions come at collegiality costs and may harm the mutual relationships of judges. -->

## The Identification-Disagreement Model of Dissenting Behavior {#identification-disagreement}
Wittig argues that the traditional theories of dissent stemming from the common law context all have limited explanatory power within the civil law context, as judges therein are deciding in a different context, bound by different procedural rules, and, thus, given differing, sometimes broader, sometimes more limited, avenues to give way to their policy preferences or strategic considerations. To alleviate these issues, Wittig introduces a non-formal model of dissenting behavior, the identification-disagreement model. The model is made up of two dimensions: the disagreement level and judges' stance and degree of self-identification of their role as a judge, termed as *norm of consensus*. Separate opinions are then "a function of a judge's identification with the norm of consensus and the level of disagreement of judges [@wittigOccurrenceSeparateOpinions2016, p. 74-75]. We base our theory on the identification-disagreement model because, institutionally, the CCC is akin to the FGCC and because the results of Wittig's empirical study corresponded to her theory.

Following the identification-disagreement model, the probability of a separate opinion depends on judges' adherence to the norm of consensus and the level of disagreement. Therefore, the first two hypotheses are as follows:

**H~3~:** *The probability of observing a separate opinion is higher for judges with low norm-identification than for judges with high norm-identification.*

**H~4~:** *The probability of observing a separate opinion is higher for cases with a higher level of disagreement potential than in cases with a lower level of disagreement potential.*

We now flesh out both of the concepts. We start with the norm of consensus and then we move on to the disagreement.

### The norm of consensus
@calderiaTimeConsensualNorms1998, p. 876-877 define a norm as "a long-run equilibrium outcome, which underpins the interaction between individuals and reflects common understandings as to what is acceptable behavior in given circumstances." The norm of consensus in turn defines the level of dissent that is acceptable at any given court [@narayanConsensualNormHigh2005; @wittigOccurrenceSeparateOpinions2016, p. 75.]. Wittig's argument is two-fold. First, in civil law traditions unlike its US counterpart, the prevailing notion of the norm of consensus is that a court should not display disagreement. Second, the extent of adherence to the norm varies among judges [@wittigOccurrenceSeparateOpinions2016, p. 75.].^[We conducted interviews with the judges of the third term of the CCC. Practically all of them more or less directly confirmed that they share the view that judges should not display dissent at a civil law court to a very varying degree.]

<!-- A dissonance between a proposed outcome for a case and any given judge's preferences are eventually bound to happen. In such a case, the judge can either express their sincere preferences by adopting a separate opinion or they can adapt their behavior according to the norm of consensus and suppress the expression of their preferences. The decision of judge faced with such a conflict whether to attach a separate opinion or whether to avert their dissent is then a function of multiple potential utilities. Firstly, the intrinsic utility is maximized whenever a judge behaves in accordance with their true values and opinions, setting aside their strategic or political considerations. Secondly, expressive utility is harnessed when one displays individuality and counters the notion of conformism. Thirdly, the reputational utility arises when one adjusts their publicly displayed preferences to the expectations of others. The decision to dissent or to avert a dissent then is a result of an weighting between costs and benefits of these three types of utilities a judge derives from adhering or not adhering to the norm of consensus. -->

How to capture the norm-identification of judges? The key dimension across which the extent of norm-identification may be theoretically expected to vary across judges is the professional background of the judges. The judges socialized within the judiciary are more likely to adhere to its values than those coming outside of it. In other words, CCC judges being appointed from the ranks of the judiciary are expected to adhere to the norm of consensus more than the rest. That applies especially strongly within the civil legal system, in which voicing a disagreement is usually prohibited within the ordinary judiciary. Bricker notes that unlike the SCOTUS, which now is almost exclusively comprised of career judges, European constitutional judges arrive to the court from diverse paths: some arrive after careers in the ordinary court system, some arrive with backgrounds in academia, and others arrive after careers in politics [@brickerBreakingPrincipleSecrecy2017]. Moreover, legal professions in civil-law systems have been comparatively much more tightly embedded into a unified state bureaucracy [@garoupaSpurredLegalTradition2020].

In Czechia, apart from the CCC and the Supreme Administrative Court^[And therein there are an extreme oddity too.], no separate opinions are even legally allowed. Therefore, the CCC judges coming out of the judiciary are socialized in an environment, in which voicing a dissent is not even allowed by law. On the other hand, lawyers or judges recruited from the ranks of the politician or scholars are not socialized from within the judiciary and are used to defending their own opinion, or at least they are not required to hide their disagreement. However, we expect that the difference between the professions should wane over time as the non-judicial judges get accustomed to the role of a CCC judge. Our hypothesis 3 suggests: 

**H~5~:** *The closer the date of the decision to the date at which the judge entered, the larger the difference between the probability of observing a separate opinion of judges coming from judiciary and of judges coming outside judiciary.*

Research into the effect of professional background of judges is no novelty. @garoupaDisagreeingPrivateDissenting2022 have found evidence that the individual decision to attach a separate opinion of the judges on the Spanish Council of State depends on the judge's professional background. The effect does not go in the direction that we would theoretically expect: only former politicians seem to be less likely to attach a separate opinion, while former judges and lawyers seem to be more likely to attach a separate opinion. While evidence in the aforecited studies concludes in the favor of presence of such an effect, in the European context, @garoupaSpurredLegalTradition2020 in their comparative study of Slovenia and Croatia found no evidence for the influence of judges' background, such as their former career or education. 

### Disagreement on the bench
A disagreement on a bench arises when the opinions on the matter diverge during a discussion and a judge has a reason to object the majority view. For example, case characteristics play an important role. Cases with more value-laden or controversial topics may give raise to more disagreement, similarly highly complex cases leave more space for disagreement. The sources of disagreement are seemingly manifold. To elucidate them theoretically, we turn to the case-space model.

The case-space model is a theoretical model developed in an attempt to model the idiosyncrasies of court decision-making such as the role of law [@landaDisagreementsCollegialCourts2007; @laxNewJudicialPolitics2011]. In the case-space model, the way to represent a case is by locating it in a n-dimensional space of of possible cases, the case-space. A case then denotes a legally relevant event that has occurred out of many that could have occurred. Put more simply, a case is a bundle of legally relevant facts. A legal rule then is a cut that divides the case-space. An individual disposition of a case is then the judgment of an individual judge of the case depending on their preference over legal rules (the cut) and the location of the case in the case-space. In other words, if the case at hand falls to one side of the judge's cut point, then they vote for certain outcome, if it falls on the other side, they vote against it.

@landaDisagreementsCollegialCourts2007 draw from the case-space model multiple theoretical sources of disagreement. The first and clear source of disagreement among judges is that about facts. Different judges may place the facts of the case in the case-space differently. There other sources of disagreement: which dimensions should be relevant under any given legal rule to determine the disposition of a case, disagreement about "thresholds" within dimensions, and a couple of more sources of disagreement, which all can be summarized as a disagreement about the legal rule. Concerning the state of the art research on the effects of disagreement on dissent, @wittigOccurrenceSeparateOpinions2016 found a positive effect of the legal complexity on the dissenting behavior of judges, similarly @brickerBreakingPrincipleSecrecy2017 found a similar effect in a study on the German, Polish, Latvian, and Slovenian constitutional courts.

To this end, we simplify the model into two similar characteristics that concern the facts and the legal rule: case complexity and case controversy. Case complexity in our understanding refers to the number of legal issues that a case has touched upon. In line with @landaDisagreementsCollegialCourts2007 we suppose that the more legal rules there is in play in any given case, the more room for disagreement about rules. We are also able to distinguish between different types of legal issues. Case controversy refers to the facts. We obtain the following two hypotheses:

**H~6a~:** *The more legally complex a case is, the higher is the probability of a judge attaching a separate opinion in that case.*

**H~6b~:** *The more controversial a case is, the higher is the probability of a judge attaching a separate opinion in that case.*

Within the CCC, we can observe a special example of circumstances giving rise to higher level of disagreement. @chmelCoOvlivnujeUstavni2021; @smekalMimopravniVlivyNa2021; @vartazaryanSitOvaAnalyza2022 found that the third term of the CCC between 2013-2023 is rather polarized and that there are two voting blocs of judges that clash against each other in the plenary decisions. We believe that such an situation is theoretically a special case of circumstances with higher level of disagreement: if the relationships between the judges are strained from the plenary proceedings, they should also carry over to the 3-member chamber hearings. If this shows to be true, it would provide a robustness check the two coalition theory as well as for the Wittig's identification-disagreement model. Therefore, our next hypothesis suggests

**H~7~:** *Having a 3-member chamber composed of members of both judicial coalitions increases the probability of a judge attaching a separate opinion in that case.*

# A brief primer on the CCC {#primer}
We now introduce the CCC, its institutional and procedural background, its powers as well as its composition. The CCC consists of fifteen judges^[I will proceed with calling the CCC judges as "judges" in the SCOTUS fashion [@boatrightHistoryMeaningUse2018] as they technically sit on the highest court in Czechia.], out of which one is the president of the CCC, two are vice presidents and twelve associate judges [following the terminology of @kosarConstitutionalCourtCzechia2020]. These fifteen judges are appointed by the president of the Czech republic upon approval of the Senate, the upper chamber of the Czech two-chamber Parliament. The judges enjoy 10 years terms with the possibility of reelection; there is no limit on the times a judge can be re-elected. The three CCC functionaries are unilaterally appointed by the Czech president.

The appointment procedure is similar to how the SCOTUS judges are appointed as the procedure lays in the hands of the president of the republic and the upper chamber. The minimal requirement for a CCC nominee are 40 years of age, a clean criminal record, a finished legal education and experience in the legal field. Other than that, the nomination is left to the consideration of the President of the Republic. After a nomination, the nominee is firstly interviewed in the constitutional law committee of the Senate, which produces an unbinding recommendation for the plenary Senate hearing. The final binding decision is then made by simple majority of the Senate plenary hearing. This procedure has lead to a situation, in which there is very little variance as to the nominating background of the judges. First, there is no nominating political party akin to the US context or the Spanish context [@hanrettyDissentIberiaIdeal2012]. Second, because the court was established in 1993 and filled within roughly a year of its establishment and because the term of the Czech president is 5 years and all the 3 Presidents, who'd finished their term at the time of writing this article, have been elected twice (for ten years it total), each president has had the chance to appoint all the fifteen members of "their" CCC. Therefore, the first term of the CCC has been termed the Václav Havel, the second the Václav Klaus and the third Miloš Zeman terms of the CCC.

Regarding the competences, the CCC is a typical Kelsenian court inspired mainly by the German Federal Constitutional Court. The CCC enjoys the power of abstract constitutional review, including constitutional amendments. The abstract review procedure is initiated by political actors (for example MPs) and usually concerns political issues. Moreover, an ordinary court can initiate a concrete review procedure, if that court reaches the conclusion that a legal norm upon which its decision depends is not compatible with the constitution. Individuals can also lodge constitutional complaints before the CCC. Lastly, the CCC can also resolve separation-of-powers disputes, it can *ex ante* review international treaties, decide on impeachment of the president of the republic, and it has additional ancillary powers [for a complete overview, see @kosarConstitutionalCourtCzechia2020].

The CCC is an example of a collegial court. From the perspective of the inner organization, the CCC can decide in four bodies: (1) individual judges in the role of judge rapporteur, (2) 3-member chambers (*senáty*), (3) the plenum (*plénum*), and (4) special disciplinary chamber. The 3-member chambers and the plenum play a crucial role. The plenum is composed of all judges, whereas the four 3-member chambers are composed of the associate judges. Neither the president of the CCC or her vice-presidents are permanents members of the 3-member chambers. Until 2016, the composition of the chambers was static. However, in 2016, a system of regular 2-yearly rotations was introduced, wherein the president of the chamber rotates to a different every 2 years. I am of the view that such a institutional change opens up potential for quasi-experimental research similar to the @gschwendAreJudgesPolitical2016 study utilizing judge absences within the 3-member chambers of the German federal constitutional court. In general, the plenum is responsible for the abstract review, whereas the 3-member chambers are responsible for the individual constitutional complaints.

In the chamber proceedings, decisions on admissibility must be unanimous, whereas decisions on merits need not be, therefore, a simple majority of two votes is necessary to pass a decision on merits. In the plenum, the general voting quorum is a simple majority and the plenum is quorate when there are ten judges present. The abstract review is one of the exceptions that sets the quorum higher, more specifically to 9 votes.

A judge rapporteur plays a crucial role [@chmelZpravodajoveSenatyVliv2017; @horenovskyProcessMakingConstitutional2015 study the large influence of the judge rapporteurs at the CCC]. Each case of the CCC gets assigned to a judge rapporteur. The assignment is regulated by a case allocation plan.^[The original term is *rozvrh práce*, which is usually translated as a *work schedule*, however, I borrow the term *case allocation plan* from @hamannGermanFederalCourts2019, p. 673.] They are tasked with drafting the opinion, about which the body then votes. The president of the CCC (in plenary cases) or the president of the chamber (in chamber cases) may re-assign a case to a different judge rapporteur if the draft opinion by the original judge rapporteur did not receive a majority of votes. Unfortunately, the CCC does not keep track of these reassignments.

The act on the CCC allows for separate opinions. They can take two forms: dissenting or concurring opinions. Each judge has the right to author a separate opinion, which then gets published with the CCC decision. It follows that not every anti-majority vote implies a separate opinion, it is up to the judges to decide whether they want to attach a separate opinion with their vote. Vice-versa, not every separate opinion implies an anti-majority vote, as the judges can attach a concurring opinion. In contrast to dissenting opinion, when a judge attaches a concurring opinion, they voted with the majority but disagree with its argumentation.^[Which makes it difficult to, for example, conduct the same point-estimation with data on dissenting behavior of judges as @hanrettyDissentIberiaIdeal2012 has done on the Portugese and Spanish Constitutional Courts.] Unfortunately, it is difficult to conduct research on the dissent aversion because the voting is kept secret.

The room for the dissenting judge and the majority to address each other differs between the two bodies. Based on our qualitative research, there is less back and forth interplay between the judges, more akin to the SCOTUS context, and most of the communication is handled remotely in the chamber proceedings, whereas the plenum meets regularly to discuss the cases in person. Despite that, procedurally speaking, the process of generating separate opinions is the same. In both cases, the rapporteurs are informed about the outcome of the vote, which is filed in the voting record. The separate opinion is then sent to the judge rapporteur before the decision is announced, as it cannot be added until after the announcement. It is important to note that judges have the possibility, not the obligation, to dissent. In other words, there is room for judges to give way to strategic considerations.

# Method {#method}
We now move onto the empirical part of our paper. First, in [section 4.1](#data-description) we describe the data we collected. Second, in [section 4.2](#operationalization) we explain how we operationalized the variables included in our models. Last, in [section 4.3](#identification) we explain our identification strategy.
t
## Data description {#data-description}
The data is based on the CCC database [@paulikCzechConstitutionalCourt2024], which contains all decisions published by the CCC since its foundation, complete text corpus as well as plenty of metadata. The whole dataset is available at [Zenodo](https://zenodo.org/records/10591110). We narrow our cases to all plenum decisions and to all 3-member chamber decisions on merits up until the end of 2022. The admissibility decisions of the 3-member chambers must be made unanimously, separate opinions are not allowed at all, concurring decisions therein are merely a rarity. Out of `r separate_opinion_table$total[[1]] + separate_opinion_table$total[[3]]` chamber admissibility and procedural decisions, only `r separate_opinion_table$n[[1]] + separate_opinion_table$n[[3]]`.^[On top of that out of the 39 concurring opinions in admissibility chamber decisions, 25 of that are a copypasta from judge Jan Filip and 6 are a copypasta from judge Josef Fiala.] In contrast to that, out of the total `r separate_opinion_table$total[[2]]` chamber decisions on merits, `r separate_opinion_table$n[[2]]` contain at least one separate opinion. Fig. \@ref(fig:separate-opinion-ratio-chamber) confirms our intuition about imbalance among the admissibility decisions. In contrast to that, the plenary decisions contain significantly larger number of separate opinions as fig. \@ref(fig:separate-opinion-ratio-plenum) shows.

```{r separate-opinion-ratio-chamber, fig.cap = "The number of the 3-member chamber decisions with at least one separate opinion."}
separate_opinion_ratio_chamber
```

Moreover, we skip the first decade of the CCC as the data on it are incomplete and inconsistent.^[For a complete overview of the extent of and details about the missingness issue, the reader can refer to @paulikCzechConstitutionalCourt2024)] For example, very few decisions contain the information on the composition of the bench in the text and many do not contain the name of the dissenting judge at all. To avoid the issue of right censored data, we limit our analysis until the end of 2022 as the CCC entered its 4th term in 2023 and started to undergo a personal overhaul that still has not been completed.

```{r separate-opinion-ratio-plenum, fig.cap = "The number of the plenary decisions with at least one separate opinion."}
separate_opinion_ratio_plenum
```

The final narrowed dataset for analysis contains `r toString(length(unique(data$doc_id)))` decisions of the CCC. Table \@ref(tab:overalltable). reveals that out of them, `r overall_table[overall_table$formation == "Chamber", "ratio_total"]` are decisions by the 3-member chambers on merits, `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "merits", "ratio_total"]` are plenum decisions on merits and the remaining `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "admissibility", "ratio_total"]` are plenum decisions on admissibility. At least one separate opinion appears in `r overall_table[overall_table$formation == "Chamber", "ratio_dissent"]` decisions out of the 3-member decisions, `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "admissibility", "ratio_dissent"]` decisions of the plenum decisions on admissibility, and `r overall_table[overall_table$formation == "Plenum" & overall_table$grounds == "merits", "ratio_dissent"]` decisions of the plenum decisions on merits.

```{r overalltable}
overall_table |>
  kable(
    caption = "Overview of the data included in the analysis.", 
    col.names = c("Formation", "Grounds", "Count", "%", "% with at least 1 SO")
  ) |>
  row_spec(0,bold=TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```


## Operationalization {#operationalization}
We now discuss how we operationalized the variables included in our hypotheses. We start with the dependent variables, then we move on to the explanatory variables and finish with control variables.

### Dependent variable: a separate opinion
We conceptualize our dependent variable as the information whether a judge attached a separate opinion to a decision or not in a certain decision.^[Unlike @wittigOccurrenceSeparateOpinions2016 or @varolEmpiricalAnalysisJudicial2017 we do not call our dependent variable a judge's vote, as that refers to a slightly different thing within the CCC context. A judge may vote against the majority opinion but since they are not mandated to write a separate opinion, these do not necessarily overlap. Similarly, a judge may vote for a disposition of a case and still attach a concurrence separate opinion.] The dependent variable of the separate-opinion level model *separate opinion* is thus a dummy variable that has two categories: either a judge did attach a separate opinion (1) or she did not (0) in any given case, the dependent variable of the case level model is also a dummy variable with two categories: (1) a separate opinion in any given case occured or (2) did not.

We do not distinguish between a concurrence and a full dissent. The reason is two-fold: theoretical and practical. Theoretically, the difference between the two lays only in the disposition of the case. The judges may equally disagree on the interpretation of legal rules, thus, in the case-space model terms [@landaDisagreementsCollegialCourts2007; @laxNewJudicialPolitics2011], the judge cut points in any given case differ even when a judge attaches "only" concurrence. The difference is that in the cases containing concurrence, the case facts may have completely accidentally fallen on the same side both of the concurring judge as well as the majority, whereas in the cases containing a dissenting opinion, the case fell in between the cut points. We do not consider this phenomenon theoretically relevant for our paper. Practically, it was impossible to distinguish between the two categories as the texts of especially the older decisions simply do not contain the information whether a separate opinion was a concurring or dissenting opinion. Attempting to impute the information by relying only on the text of the separate opinion would at best produce inaccuracy.

### Explanatory variables
#### Disagreement potential {-}
Based on our theory we expect the potential for disagreement to be captured by two characteristics of any given case: (1) its legal complexity and (2) its controversy or in terms of @epsteinMeasuringIssueSalience2000 issue salience.

##### Complexity {-}
As @corleyPuzzleUnanimityConsensus2013 a define legal complexity as the number of legal issues a case has to address. True to the @corleyPuzzleUnanimityConsensus2013 study [and others which have operationalized legal complexity similarly such as @spriggsExplainingOverrulingSupreme2001], our operationalization of legal complexity relies on the assumption that the more legal issues there are in any given case, the higher the number of other laws and caselaw that the court has relied on in any given decision. A case with higher legal complexity will have relied on more legal provisions or court decisions than a relatively simple case. 

Because only the constitutional-level acts (which includes the Constitution, the fundamental human rights, as well as the EU law in some cases) serve as the benchmark for the review of laws and individual complaints,^[See Article 87 of the Czech Constitution] only constitutional-level sources of law are relevant. Moreover, because phrasing of the Czech Constitution and Charter articles is very vague, even a relatively complex case may concern just one or two articles of the Constitution. As a result, CCC decisions mostly rely on its own caselaw. Thus, we capture the complexity of a case in the variable *caselaw* as a number of other CCC decisions mentioned in the text of the decision.

##### Controversy/Issue salience {-}
In a similar vein, certain typically value-laden issues may generate more disagreement even if they raise only one or few legal questions. @epsteinMeasuringIssueSalience2000 distinguish between two types of issue salience. Retrospective issue salience is how a particular issue is viewed in hindsight, regardless of whether the actors at the time regarded it so. Contemporaneous salience than captures the stance of the relevant actors, the CCC judges in our cases, on the issue. While we are interested in the latter, it is not obvious as to how exactly can such an information be captured. @epsteinMeasuringIssueSalience2000 reject all the typically employed proxies for issue salience such as the number of articles in law review that a case generated or the number of citations a case generated. Instead, they propose a novel unbiased transparent measure of issue salience: whether a SCOTUS decision made it to the front-page of the New York Times newspaper.^[The question is to what extent does that not also reflect the retrospective issue salience.] Unfortunately, practically all of the measures are hard to transfer to the Czechia context. We cannot exactly pinpoint a similar newspaper as New York Times in Czechia. Therefore, we need to rely on another measure of case controversy that is already discernible at the time of the decision.

@wittigOccurrenceSeparateOpinions2016 selected a number of subject matters she deemed as particularly controversial. In the CCC context, there are a number of issue that have been deemed as controversial by the Czech legal scholarship. Typically, the restitution cases or cases concerning fundamental human rights have been coined as rather controversial. The CCC database contains a table *subject_matter*, which contains the subject matter of any given case (a discrimination case, a separation of powers case and alike) and information on the area of constitutional law that the case pertains (the right to fair trial, the right of freedom of speech). We coded the subject matter as controversial when they concern the fundamental issues of state organization in the Czech context or when they concern ethical issues. Regarding the former, expropriation [@radvanExpropriationWiderPerspective2018], restitutions, the separation of state and church, as well as social rights have played an especially sensitive role after the transition from the Soviet regime [@bobekKomunistickePravoCeskoslovensku2009]. Moreover, ethical issues such as discrimination [@havelkovaGenderEqualityLaw2017], same sex marriage, transgender rights, or human dignity have recently enjoyed greater attention in the society. The former group of topics is split along the left-right axis, whereas the later group of topics is split along the progressive-conservative axis. The dummy variable *controversial* was coded as 1 if a decision pertained any of the following subject matters: discrimination, expropriation, restitution, sexual orientation, fundamental human rights, social and cultural rights, the right of property, the freedom of speech, and the separation between the church and state. Topics such as fair trial, civil law issues, or procedural issues have been coded as uncontroversial.

#### Norm-identification {-}
Secondly, our theory generated a second expectation: a relationship between adherence to the norm of consensus and the likelihood of a separate opinion. The adherence to the norm of consensus depends on self-identification and on self-perception of the judges: "being part of the system of the judiciary influences identification with the organization's goals and it shapes considerably the judges' perception of how to behave." [@wittigOccurrenceSeparateOpinions2016, p. 100].

We operationalize norm identification as the last profession before they entered the CCC interacting with the number of months since their appointment. Viewed this way, we presume that the last profession that a judge had held before entering an office had the highest impact on their self-identification as a judge. Overtime we expect the effect to wane off.

```{r profession-distribution, fig.cap = "Distribution of the professions in our final dataset. Scholars, lawyers, and politicians are flattened into \\textit{outside} value of the \\textit{profession} variable."}
professions
```

In the CCC database, the variable *judge_profession* contains the information on the judges' previous career choices and can take up the values of judge, scholar, politician, or lawyer. The table  \@ref(fig:profession-distribution) reveals the distribution of the professions. For our purposes, we flatten the original variable into 2 levels. The variable *profession* is coded either as "within" if the judge previously served as a judge or "outside" if they came from the other three professions. Therefore, the final dataset contains `r n_within` judges coming out of the judiciary and `r n_outside` coming outside of the judiciary.

#### Time in office {-}
The effects of the CCC judges' profession should according to our theoretical expectations vary depending on the time the judges have spent in their office. The closer to them taking up the office, the effects of their preceding profession should be more pronounced. Similarly, according to the collegiality costs hypothesis, the dissenting behavior of judges should vary across time. Namely, judges are expected to dissent less likely at the beginning of their terms, as the collegiality costs are higher, and to dissent more likely at the end of their terms, as the collegiality costs of the disagreement are lower. We capture the *time in office* variables as the number of months a judge that have elapsed since the judge's appointment. We note that since reelection is allowed and that `r reelected_judges` judges in our dataset have been once re-elected, the count starts already at the time of the first appointment as the socialization of a reelected judge occurs from the beginning of their first mandate. Similarly, to account for hypothesis 1, we also include the variable *time until end* that captures the number of months until the end of the term of the judge at the time of the decision. There is one substantial difference between the two variables. The *time until end* variable does not take into account the potential for reelection as the reelection takes place only shortly before the start of the term and the reelected judges cannot during their terms foresee their future reelection. 

#### Workload {-}
Similarly to the study of @brekkeThatOrderHow2023 on workload at the CJEU, we operationalize *workload* as the number of pending unfinished cases that any given judge has in the moment of any given decision as a judge rapporteur. We used the information on the composition of the bench in any given decision from the CCC database. We then calculated the number of unfinished cases each judge had at the time of any given decision as a judge rapporteur using the date of submission and of decision of a case. The mean number of pending cases of any judge as a judge rapporteur is `r descr_workload$numeric.mean[1]`, with quite a large standard deviation of `r descr_workload$numeric.sd[1]`. The highest number of pending cases in the dataset is `r descr_workload$numeric.p100[1]` in the case of the judge Radovan Suchánek, whose work ethic has already become a topic of scholarly discussion [@chmelCoOvlivnujeUstavni2021].

<!-- One looming issue arises that needs to be addressed and may introduce bias. Not every unfinished pending case is the same. Different judges may cope with the increasing workload differently. Czech scholars have argued that deciding cases on admissibility, which are typically shorter and require a rather concise argumentation, may be such a coping mechanism of overburdened judges: it allows them to expedite decisions faster and with less effort. Thus, the same number of unfinished pending cases of a judge that chooses the strategy of offloading their workload burden into decisions on admissibility may not carry the same workload as with a judge that opts for decisions on merits more often. As a solution, we add this ratio in the name of *ratio_admissibility* into our model as a control variable. -->

#### Mixed coalition
Lastly, we include a dummy variable mixed coalitions, which takes up the value 1 when two conditions are met: (1) the decision was made in a 3-member chamber and (2) the composition of the chamber was made up of judges from both coalitions.

Lastly, tables \@ref(tab:dv-overview-numeric) and \@ref(tab:dv-overview-factor) present summary statistics for all dependent variables included in the models.

```{r dv-overview-numeric}
dependent_variables_overview_numeric |> 
  mutate(skim_variable = c("# of Citations", "Workload", "Time in Office", "Time until End")) |>
  kable(
    caption = "Summary statistics for the numeric dependent variables.", 
    col.names = c("", "Mean", "SD", "P0", "P25", "P50", "P75", "P100"),
    digits = 2
  )
```


```{r dv-overview-factor}
dependent_variables_overview_factor |>
  select(-group) |>
  kable(caption = "Summary statistics for the categorical dependent variables. The large number of outside and within judiciary values is due to the fact that each judge has decided at least hundreds of cases. For an overview of the number of judges with said background, see the section on operationalization of the variable profession.",
        col.names = c("", "Count")) |>
   pack_rows(index = table(fct_inorder(dependent_variables_overview_factor$group)))
```


## Identification Strategy {#identification}
The hypotheses are tested by fitting a generalized linear model estimating the probability of a judge attaching a separate opinion with the dependent variable following a Bernoulli distribution. The data at hand are a time-series data with multiple sources of unobserved heterogeneity that we address by including fixed effects. 

``` {r model-selection}
model_selection |>
  select(-group) |>
  kable(caption = "Descriptive statistics for the fixed effects clusters.",
        col.names = c("Type", "Value")) |>
  pack_rows(index = table(fct_inorder(model_selection$group)))
```

We include chamber fixed effects $\alpha_{c}$ that control for time-invariant chamber-level factors. First, the different dynamics between the 15-member plenum  and the 3-member chambers could produce bias. The more controversial cases or more legally complex cases are typically assigned to the plenum and the occurrence of separate opinions is higher in the plenary decisions. Moreover, the behavior of the 3-member chambers varies as well across different chambers. Secondly, the political heterogeneity of judges is a typical source of bias in judicial behavior studies. Practically, we can neither directly estimate it using the voting behavior of justices [@segalIdeologicalValuesVotes1995], because the votes of the CCC judges are not published, nor can we rely on the separate opinions themselves to produce point estimates of the political positions of judges, because unlike the @hanrettyDissentIberiaIdeal2012 a vote against the majority resolution does not necessitate a separate opinion and because that would introduce endogeneity into our model as it would be determined and measured through the dependent variable, the dissenting behavior. Theoretically, we are doubtful that the traditional liberal-conservative split plays that big of a role at the CCC. As explained in [section 3](#primer), the CCC judges are nominated and appointed by the President of the Republic, who traditionally plays an apolitical role in the Czech political system, the CCC judges are not political candidates of Czech political parties [as in the case of the US or of Spain @hanrettyDissentIberiaIdeal2012]. @sipulovaCzechConstitutionalCourt2018 has argued that the CCC has been relative spared of political influence. Despite that including chamber fixed effects should absorb any potential political heterogeneity of judges sitting in one chamber as it varies across chambers and it should be constant across time. In our model, the variable *formation* captures the information on the 3-member chambers, whose composition is relative unstable especially after the introduction of system of rotations of judges in 2016, therefore the information is coded as the exact composition of any given 3-member chamber, and the information whether the case was decided by the plenum, whose composition is relatively stable within the two terms.

As a robustness check, $\delta_j$ judge fixed effects that control for time invariant judge-level factors that are potentially correlated with both the independent as well as dependent variables of interest. For example, the judges' philosophy may influence both their stance towards separate opinions and their career choices or the way the view controversial cases. The drawback of including the judge fixed effects is that they absorb all time invariant judge-level factors [@wooldridgeIntroductoryEconometricsModern2019, p. 463-464], including the professional background of the CCC judges. Therefore, we include a model both with and without judge fixed effects to learn about the effects of professional background as well as to compare the estimated coefficients of other covariates.

Lastly, we include year fixed effects $\gamma_t$. The included data spans two terms of the CCC. Its second decade roughly between 2003 and 2013 and its third decade between 2014 and 2023. The composition of the CCC has completely changed between the two terms and so has, we expect, changed the behavior of the court as such. Moreover, the political and legal landscape has also developed: both parliamentary (both upper and lower chambers) and presidential elections have taken place, legal reforms have been passed. Following the findings of @garoupaSpurredLegalTradition2020, the political context may be correlated both with the willingness of the judges to dissent as well as, for example, their caseload as the political actors can lodge constitutional complaints.

Therefore, our final model is the result of the following function

$$ Pr(separate\ opinion) = \frac{1}{1 + e^{-(\sum{\beta_{i}x_{itpj}} + \alpha_p + \delta_j + \gamma_t)}}$$
, where $\alpha_p$ are the chamber for panel *p*, $\delta_j$ are the judge fixed effects, $\gamma_t$ are time fixed effects for year *t*, and the covariates in the $\sum{\beta_{i}x_{itpj}}$ are:

$$\beta_1n\_references + \beta_2profession + \beta_3workload + \beta_4controversial + $$ $$\beta_5time\_in\_office + \beta_6profession*time\_in\_office + \beta_7time\ until\ end + \beta_8admissibility$$

We also build a second model for the purpose of our 4th hypothesis. To this end, because we are interested in information on the chamber level, our unit of observation is flattened to a decision level. Our dependent variable is the information whether a separate opinion occurred or not in that decision as in the 3-member chamber decisions, the decisions have to be either made unanimously or with a 2 votes majority. Our explanatory variable is the information whether the chamber was composed fully of either coalition or whether the composition was mixed. Because case assignment^[The cases are assigned randomly to judges rapporteur based on the first letter of their surname.] as well as the assignment of judges to chambers is as good as random, and thus there is little room for confounding, we build only a very simple logistic model with data filtered only to include 3-member chamber decisions within the 3rd term of the CCC. The model in itself is more of a robustness check of the conclusions of Czech legal scholarship rather than a full fledged model built for causal inference.

# Results and discussion {#empirics}
The results summarised in tab. \@ref(tab:model-summary) reveal quite an interesting trend at the CCC. We included the model under two specification. When the judge-fixed effects are included, they also absorb the effect of the profession variable as that remains time-invariant within the individual judges. The difference between the estimates is down to the fact that the model with judge fixed effects considers only the within variation among individual judges. We take the estimate for the interaction term between *time in office* and *profession* variables from the first model and the remaining estimates from the full model with all fixed effects.

```{r model-summary, warning = FALSE, echo = FALSE}
modelsummary::modelsummary(list(
  "(1)" = model_fe_year_chamber,
  "(2)" = model_fe_year_chamber_judge),
  output = "kableExtra",
  estimate = "{estimate}{stars}",
  statistic = "({std.error})",
  exponentiate = TRUE,
  stars = TRUE,
  title = "Results from the Logit Model. The coefficients have been exponentiated.",
  coef_rename = c("Citations", "From Within Judiciary", "Time in Office", "Time until Term End", "Controversial", "Workload", "Admissibility", "From Within Judiciary X Time in Office")) |>
  kable_styling(latex_options = "HOLD_position")
```

Moving on to the interpretation of the results, our results do support the first hypothesis. According to the *Time until Term End* estimate, for each month that the decision is decided further from the judge's end of term, the odds of them attaching a separate opinion decrease by `r round(-100*(model_fe_year_chamber_judge |> tidy() |> filter(term == "time_until_end") |> pluck("estimate", 1) |> exp() - 1), 2)` %. Therefore, in line with our collegiality costs theory, the closer the decision is to an end of term, the more likely the judges are to dissent, or, put differently, less likely to avert a dissent.

Our results support the second hypothesis. Higher *workload* of a judge decreases their probability of attaching a separate opinion. The odds of a judge attaching a separate opinion to a decision decrease by `r round(-100*(model_fe_year_chamber_judge |> tidy() |> filter(term == "workload") |> pluck("estimate", 1) |> exp() - 1), 2)` % for each case that they have unfinished at the time of the decision. Once again, the effect is not negligible given that the mean of unfinished cases is `r round(mean(data$workload), 1)` with a standard deviation of `r round(sd(data$workload), 1)`. The results reveal that judges of the CCC give way to strategical considerations regarding their leisure. If they're overburdened with work, they reduce the load elsewhere, namely in the additional burden of writing separate opinions.

Unlike at the GFCC, we do not find evidence for the norm-consensus operating according to our theoretical expectations generated by the third and fifth hypotheses. For each month that passes since a judge gets appointed, the odds of them attaching a separate opinion decreases by `r round(-100*((model_fe_year_chamber |> tidy() |> filter(term == "judge_professionwithin:time_in_office") |> pluck("estimate", 1) + model_fe_year_chamber |> tidy() |> filter(term == "time_in_office") |> pluck("estimate", 1)) |> exp() - 1), 2)` % for the judges that had already served at a court prior to their appointment ($\beta_{Profession_{within}*Time\ in\ Office} + \beta_{Time\ in\ Office}$) and by `r round(100*((model_fe_year_chamber |> tidy() |> filter(term == "time_in_office") |> pluck("estimate", 1)) |> exp() - 1), 2)` % for the judges coming outside judiciary ($\beta_{Time\ in\ Office}$), a difference that is not statistically significant in the model without the judge fixed effects.^[As otherwise the time-invariant profession of the judge before their appointment are absorbed by the judge fixed effects.] Fig. \@ref(fig:predicted-probabilities) plots the predicted probability conditional upon the professional background and the time in the office.^[Implemented using the R package Marginal Effects [@arel-bundockHowIntepretStatistical]]. The figure reveals that although the behavior of judges with differing professional background converges over time, the difference is insignificant. That does not however lead us to reject the identification-disagreement as a theoretical model. We believe it to be a clear theoretical model that can guide further empirical inquiries. It may just be the case that the proxy of the judges' profession may not be wholly accurate or that even if it was there may simply be currently no norm of consensus operating at the CCC.^[We have conducted interviews with the CCC judges and they have revealed exactly that: there is little to none agreement as to the role and perception of the CCC.]

In contrast to that, we find a clear support for our fourth hypothesis and sixth hypotheses: all the disagreement potential variables are positively and significantly correlated with the likelihood of a judge attaching a separate opinion. No matter the model specification, we always found a clear statistical, substantive and positive relationship between the probability of a judge attaching a separate opinion and the complexity of a case. For each further CCC decision that gets cited according to the *Citations* estimate, the odds of a judge attaching a separate opinion increases by `r round(100*(model_fe_year_chamber_judge |> tidy() |> filter(term == "n_citations") |> pluck("estimate", 1) |> exp() - 1), 2)` %. Given that the average number of citations per decision is `r round(mean(data$n_citations), 1)` with a standard deviation of `r round(sd(data$n_citations), 1)`, the increase is not negligible.

```{r predicted-probabilities, fig.cap = "Predicted probability of a separate opinion occuring conditional on the previous profession of a judge and the number of months that has elapsed since their appointment."}
marginal_probabilities +
  ggtitle(NULL)
```

The same applies to the controversial value-laden topics. The odds of a judge attaching a separate opinion increases by staggering `r round(100*(model_fe_year_chamber_judge |> tidy() |> filter(term == "controversial1") |> pluck("estimate", 1) |> exp() - 1), 2)` % in decisions concerning a controversial topic in comparison to uncontroversial cases all other variables being held fixed . Because we are aware of the potential issue with "arbitrarily" selecting a couple of potential subject matters, we decided to test the robustness of our *controversial* variable. To do so, we conducted a placebo test. We sampled 10 placebo samples of different subject matters (n = `r length(controversial_topics)`) and fitted a mixed-effects model with the same dependent and independent variables, with the same fixed effects, the only difference being the *controversial* variable being replaced by the placebo. Heeding the @hartmanEquivalenceApproachBalance2018 paper on placebo tests, we than run an equivalence test with the following hypotheses:

$$H_{0_{Equivalence}}: \beta_{controversial} - 0.36SE_{\beta_{controversial}} < \beta_{placebo} < \beta_{controversial} + 0.36SE_{\beta_{controversial}}$$
$$H_{0_{Non-Superiority}}: \beta_{placebo} \geq \beta_{controversial} + 0.36SE_{\beta_{controversial}}$$
$$H_{0_{Non-Inferiority}}: \beta_{placebo} \leq \beta_{controversial} - 0.36SE_{\beta_{controversial}}$$


We run the placebo test 10 times, tab. \@ref(tab:placebo-test) displays the result.  None of the placebo effects passed the equivalence test, i.e. the null hypothesis of the effect of the placebo falling within the equivalence range set by the $H_0$ may be rejected. Moreover, all the placebo test fail to reject the test of non-inferiority, i.e. we can reasonably conclude that the effects of the placebo are significantly lower than of the $\beta_controversial$.

``` {r placebo-test}
placebo_models |>
  mutate(term = paste0("Placebo", 1:10)) |>
  kable(caption = "The result of 10 models fitted with different placebo sample from the subject matter variable. The models have been fitted under the same specification as the full model, we are only including the estimates for the placebo variable for the sake of clarity.",
        col.names = c("", "Estimate", "p(non-Zero)", "p(Equivalence)", "p(Non-Superiority)", "p(Non-Inferiority)"),
        digits = 2) |>
  kable_styling(latex_options = "hold_position")
```

In line with our seventh hypothesis based on the theoretical expectations generated by the Czech legal scholarship as well as the disagreement potential, we find that a separate opinion is more likely to occur in the 3-member chamber cases, whose composition is mixed by members of both voting blocs from the plenary proceedings. Tab. \@ref(tab:model-coalitions) reveals that the size of the effect is not negligible. The odds of a separate opinion occurring increase by `r round(100*(model_coalitions |> tidy() |> filter(term == "coalitionmixed_coal") |> pluck("estimate", 1) |> exp() - 1), 2)` %. While we are aware that our model is not thorough enough for any (causal) inference, it is rather a form of robustness check to the literature generated by the Czech legal scholarship that we'd summarized in the theoretical section.

```{r model-coalitions}
modelsummary::modelsummary(model_coalitions,
                           estimate = "{estimate}{stars}",
                           statistic = "({std.error})",
                           output = "kableExtra",
                           exponentiate = TRUE,
                           stars = TRUE,
                           title = "Results from the Logit Model on Coalitions",
                           coef_rename = c("Intercept", "Mixed Composition")) |>
  kable_styling(latex_options = "hold_position")
```

# Conclusion {#conclusion}
We examined the factors influencing dissenting opinions by judges at the CCC. Our findings do not support the existence of a strong norm of consensus at the CCC, unlike the GFCC. However, the complexity of a case, measured by the of citations to CCC cases is positively correlated with the probability of a dissenting opinion. Similarly, cases concerning controversial topics are more likely to generate dissents. A placebo test strengthens this finding by demonstrating that randomly chosen, non-controversial topics have minimal impact on dissenting behavior. The effect of voting blocs in the plenary proceedings seems to carry over to the 3-member chamber proceedings. The results also reveal that judges make strategic considerations. When facing a high workload, judges are less likely to write separate opinions, suggesting they prioritize workload management. Lastly, CCC judges seem to take into account collegiality costs as their are more likely to dissent at the end of their terms than at their start.

To conclude, we attach an important disclaimer. With these results, we do not pretend to make causal claims about the features of the CCC.  Although we have tried our best to control for as many potential sources of bias, our study nonetheless relies on observational data and the proxy variables we have employed are far from perfectly capturing the underlying feature. Despite that we believe to have contributed to the empirical legal research on dissenting behavior, even if with descriptive statistics at best. We have shed some light into the functioning of the CCC, which so far not been a subject of thorough empirical research. Due to its similarity with other courts, our results are not necessarily idiosyncratic to the CCC.

\vspace{30pt}

# References {-}

<div id="refs"></div>